---
title: "Tracking QC Analysis"
author: "Jordan Albrecht, Noemi Andor"
format: 
  html:
    self-contained: true
    embed-resources: true
    code-fold: true
    code-tools: true
    df-print: paged
editor: source
theme: sandstone
whitespace: large
---

## Data

**Plate map**: An xlsx with variables Row, Ploidy, and SUM159 of varying concentrations of treatment. This study has four replicates of 2N and 4N SUM159.

```{r}
#| label: plate-map
#| echo: false
#| warning: false
library(readxl)
library(matlab)
library(lubridate)
# root="~/Projects/PMO/HighPloidy_DoubleEdgedSword/data/BreastCancerCLs/SUM159/K00_GemcitabineExposure_033023"
# path2repo="~/Repositories/Gemcitabine-model"

root <- "C:/Users/80027908/Desktop/K00_GemcitabineExposure_033023"
path2repo="~/Gemcitabine-model"

Gemcitabine_PlateMap_20230327 <- read_excel(paste0(root, filesep, "Gemcitabine_PlateMap_20230327.xlsx"))
Gemcitabine_PlateMap_20230327

```

**Ilastik Output**: A csv with IDs etc

**Classifier Output**: 

## Tracking Results Analysis

Call the Ilastik tracking file.

```{r}
#| label: loading
#| warning: false
#| message: false

#Requisite loading of packages
library(tsfeatures)
library(matlab)
library(xts)
library(celltrackR)
library(tidyverse)
library(patchwork)
library(umap)
library(corrr)
library(ggcorrplot)
library(FactoMineR)
library(factoextra)
library(Metrics)

#Set to folder containing images, tracking results, plate table...
setwd("C:/Users/80027908/Desktop/K00_GemcitabineExposure_033023/H02_20230831_CellTracking_Ilastik")

#Import hand annotations
Book1 <- read_excel("C:/Users/80027908/Desktop/K00_GemcitabineExposure_033023/Book1.xlsx")
Book2 <- read_excel("C:/Users/80027908/Desktop/K00_GemcitabineExposure_033023/Book2.xlsx")

#Retired file pathways
# dat1 <- read.csv("C:/Users/80027908/Desktop/K00_GemcitabineExposure_033023/H02_20230831_CellTracking_Ilastik/E2_1_T=50_App=1001_Disapp=1.5_CSV-Table.tiff.csv")
# dat2 <- read.csv("C:/Users/80027908/Desktop/K00_GemcitabineExposure_033023/H02_20230831_CellTracking_Ilastik/E2_1_T=50_App=1001_Disapp=1001_CSV-Table.tiff.csv")
# dat1 <- mutate(dat1, ID = "E2_1_T=50_App=1001_Disapp=1.5")
# dat2 <- mutate(dat2, ID = "E2_1_T=50_App=1001_Disapp=1001")

# Open the two tracking result tables for comparison
dat1 <- read.csv("C:/Users/80027908/Desktop/K00_GemcitabineExposure_033023/Tracking_12-15/E2_1_CSV-Table_P2.h5.csv")
dat2 <- read.csv("C:/Users/80027908/Desktop/K00_GemcitabineExposure_033023/Tracking_12-15/E2_1_CSV-Table_P3.h5.csv")
dat1 <- mutate(dat1, ID = "E2_1_P2")
dat2 <- mutate(dat2, ID = "E2_1_P3")

# Read in the classifier file
seg=read.csv("M:/lab/lab_crd/HighPloidy_CostBenefits/data/BreastCancerCellLines/SUM-159/K00_GemcitabineExposure_033023/I01_20231219_HALOCellClassification/E2_1_Total_Object_Results.csv")
# seg=read.csv("C:/Users/80027908/Desktop/K00_GemcitabineExposure_033023/B01_20230407_Incucyte_Images_woGFP_Analysis_QI_Core/Results_Classifier_2/E_objectdata.csv")

well="E2_1_"
```

Create the tracking plot.

```{r}
#| label: plot-traj
#| fig-cap: Trajectories of Set 1.
#| warning: false
#| message: false

#Get the dimensions of the frame of motion for every cell?
ii=match(c("frame","Bounding_Box_Maximum_0","Bounding_Box_Maximum_1"),colnames(dat1))
jj=match(c("frame","Bounding_Box_Maximum_0","Bounding_Box_Maximum_1"),colnames(dat2))

colnames(dat1)[ii]=c("t","x","y")
colnames(dat2)[jj]=c("t","x","y")
#List dfs of cells in well
la=sapply(unique(dat1$trackId), function(x) dat1[dat1$trackId==x,c("t","x","y","Object_Area_0")], simplify = F)
la2=sapply(unique(dat2$trackId), function(x) dat2[dat2$trackId==x,c("t","x","y","Object_Area_0")], simplify = F)
#Calls cells as their trackID
names(la)=as.character(unique(dat1$trackId))
names(la2)=as.character(unique(dat2$trackId))
la=la[!names(la) %in% c("-1")]
la2=la2[!names(la2) %in% c("-1")]
plot(as.tracks(la))
```
```{r}
#| warning: false
#| message: false
#| label: fig-2
#| fig-cap: Trajectories of Set 2.

plot(as.tracks(la2))
```

## Merge with classifier results

**1.** Read in classifier data

```{r}
#Pull in the data for the well of interest
seg=seg[grep(well,seg$Image.Location),]
seg$time=sapply(strsplit(seg$Image.Location,"_"), function(x) x[length(x)])
seg$time=gsub(".tif","",seg$time)
```

**2.** Change times to frames

```{r}
#Replace time with frame
map=0:(length(unique(seg$time))-1)
names(map)=sort(unique(seg$time))
seg$t=map[seg$time]
seg=seg[order(seg$t),]

#Check changes
boxplot(seg$t~seg$time, horizontal = T,las=2,ylab="",cex.axis=0.54)
```

**3.** Compare x and y values

```{r}
#Pull x and y columns
ii=match(c("XMax","YMax"),colnames(seg))
colnames(seg)[ii]=c("x","y")
coi1=intersect(colnames(seg),colnames(dat1))
coi2=intersect(colnames(seg),colnames(dat2))
ii=sapply(map, function(t) list(seg=which(seg$t==t), dat1=which(dat1$t==t)), simplify = F)
jj=sapply(map, function(t) list(seg=which(seg$t==t), dat2=which(dat2$t==t)), simplify = F)
ii=ii[sapply(ii, function(x) min(length(x[[1]]),length(x[[2]])))>0]
jj=jj[sapply(jj, function(x) min(length(x[[1]]),length(x[[2]])))>0]
d1=sapply(ii, function(t) flexclust::dist2(seg[t$seg,coi1,drop=F],dat1[t$dat1,coi1,drop=F]))
d2=sapply(jj, function(t) flexclust::dist2(seg[t$seg,coi2,drop=F],dat2[t$dat2,coi2,drop=F]))
```

**4.** Assign tracked cells to segmented cells.

```{r}
#Combining dfs using names and x values ?
seg_matched=sapply(names(d1), function(x) seg[ii[[x]]$seg,][apply(d1[[x]],2,which.min),],simplify = F)
merged1=sapply(names(d1), function(x) cbind(dat1[ii[[x]]$dat1,],  seg_matched[[x]][,c("time","Classifier.Phenotype")]), simplify = F)
merged_B1=merged1
merged1=do.call(rbind, merged1)
merged1=merged1[order(merged1$t),];

seg_matched=sapply(names(d2), function(x) seg[jj[[x]]$seg,][apply(d2[[x]],2,which.min),],simplify = F)
merged2=sapply(names(d2), function(x) cbind(dat2[jj[[x]]$dat2,], seg_matched[[x]][,c("time","Classifier.Phenotype")]), simplify = F)
merged_B2=merged2
merged2=do.call(rbind, merged2)
merged2=merged2[order(merged2$t),];
```

**5.** Find the center of each cell.

```{r}
merged1$y=(merged1$Bounding_Box_Minimum_1+merged1$y)/2
merged1$x=(merged1$Bounding_Box_Minimum_0+merged1$x)/2
merged1$y=-merged1$y
merged1$y=merged1$y-min(merged1$y)
merged2$y=(merged2$Bounding_Box_Minimum_1+merged2$y)/2
merged2$x=(merged2$Bounding_Box_Minimum_0+merged2$x)/2
merged2$y=-merged2$y
merged2$y=merged2$y-min(merged2$y)

#View changes
select(merged1, labelimageId, mergerLabelId, Object_Center_0, Object_Center_1, x, y, Bounding_Box_Minimum_0, Bounding_Box_Minimum_1, Center_of_the_object_0, Center_of_the_object_1)
select(merged2, labelimageId, mergerLabelId, Object_Center_0, Object_Center_1, x, y, Bounding_Box_Minimum_0, Bounding_Box_Minimum_1, Center_of_the_object_0, Center_of_the_object_1)
```

The 'x' and 'y' values in the merged table were brought in from the classifier, and the 'Bounding_Box' values are from the tracking set ?

**6.** Make Dead, Transitional, and Alive statuses into a range from -1 to 1.

```{r}
merged1$Class=1
merged1$Class[merged1$Classifier.Phenotype=="Transitional"]=0
merged1$Class[merged1$Classifier.Phenotype=="Dead"]=-1
merged2$Class=1
merged2$Class[merged2$Classifier.Phenotype=="Transitional"]=0
merged2$Class[merged2$Classifier.Phenotype=="Dead"]=-1
```

## Existing Quality Control

**1.** See if number of tracked cells is same as number of segmented cells.

```{r}
#| label: fig-3
#| fig-cap: Pearson correlation for E2_1_T=50_App=1001_Disapp=1.5
#| warning: false

#Set 1
sapply(ii, function(x) c(length(x$dat1),length(x$seg)))

#Pull tracked cells
la1=sapply(unique(merged1$trackId), function(x) merged1[merged1$trackId==x,c("t","x","y","Object_Area_0","time","Classifier.Phenotype","Class")], simplify = F)
names(la1)=as.character(unique(merged1$trackId))
la1=la1[!names(la1) %in% c("-1")]
la_1=la1[sapply(la1,nrow)>=10]

#Uhhh
# la_1=sapply(la_1, function(x) x[x$t<(24*10)/4,],simplify = F)
# la_1=la_1[sapply(la_1,nrow)>=10 & sapply(la_1, function(x) min(x$t))<100]
te1=sapply(la_1, function(x) cor.test(x$t,x$Class)$estimate)
hist(te1, xlab="Pearson correlation(time, cell state)",main="Correlation between time and assigned class")
mtext("Dead cells should not come back to life (corr should be negative for dying cells)")
mtext("corr should be near-zero for cells that never die",line = 3)
la_1=la_1[is.na(te1) | te1<0.05];
la_1=la_1[!sapply(la_1,is.null)];
print(paste(length(la_1)/length(te1),"cells included (which do not come back to live from dead)!"))
```

```{r}
#| label: fig-4
#| fig-cap: Pearson correlation for E2_1_T=50_App=1001_Disapp=1001
#| warning: false

#Set 2
sapply(jj, function(x) c(length(x$dat2),length(x$seg)))

#Pull tracked cells
la2=sapply(unique(merged2$trackId), function(x) merged2[merged2$trackId==x,c("t","x","y","Object_Area_0","time","Classifier.Phenotype","Class")], simplify = F)
names(la2)=as.character(unique(merged2$trackId))
la2=la2[!names(la2) %in% c("-1")]

la_2=sapply(la2, function(x) x[x$t<(24*10)/4,],simplify = F)
la_2=la_2[sapply(la_2,nrow)>=10 & sapply(la_2, function(x) min(x$t))<100]
te2=sapply(la_2, function(x) cor.test(x$t,x$Class)$estimate)
hist(te2, xlab="Pearson correlation(time, cell state)",main="Correlation between time and assigned class")
mtext("Dead cells should not come back to life (corr should be negative for dying cells)")
mtext("corr should be near-zero for cells that never die",line = 3)
la_2=la_2[is.na(te2) | te2<0.05];
la_2=la_2[!sapply(la_2,is.null)];
print(paste(length(la_2)/length(te2),"cells included (which do not come back to live from dead)!"))
```

**2.** Visualize change in size over time. Removed due to disturbingly long null output.

**3.** Visualize one track at a time on raw images. This cannot be run due to missing files.

## New Quality Control

**1.** Check size of parent and daughters for cell division events.
```{r}
#| label: plot-ratio
#| fig-cap: Ratio of parent cell size to average daughter size.
#| warning: false

merged=list(merged1=merged1, merged2=merged2)
for(data in names(merged)) {
  df <- get(data)
  #Find a list of all the parents
  parents_list <- unique(df$parentTrackId)
  parents_list <- parents_list[!parents_list %in% 0]
  #Find all of the data points for each parent
  parents_df <- df %>% subset(trackId %in% parents_list)
  #Find all of the daughters of each parent in the list
  daughters_df <- df %>% subset(parentTrackId %in% parents_list)
  assign(paste("daughters", data, sep = "_") , daughters_df)
  #Finding only the areas of parents at latest time point (right before division)
  parents_max <- data.frame()
  for(i in parents_list){
    p <- filter(parents_df, trackId == i)
    q <- max(p$t)
    r <- filter(p, t == q)
    parents_max <- rbind(parents_max, r)
  }
  assign(paste("parentsmax", data, sep = "_") , parents_max)
  #Finding the average area of each daughter
  daughters_means <- data.frame()
  for(i in parents_list){
    p <- filter(daughters_df, parentTrackId == i)
    daughtersArea <- mean(p$Object_Area_0)
    r <- cbind(trackId = i, daughtersArea)
    daughters_means <- rbind(daughters_means, r)
  }
  #Visualize
  ratios <- left_join(parents_max, daughters_means) %>% 
    mutate(ratio = Object_Area_0/daughtersArea)
  print(ggplot(ratios, aes(x = ratio)) + 
    geom_histogram(bins = 150) + 
    geom_vline(aes(xintercept=mean(ratio)), 
               color="blue", 
               linetype="dashed", 
               size=1) + 
    theme_light() + 
    ggtitle(paste0("Parent versus Average Daughter Cell Size for ", df[1,63])))
}
```
An average ratio of 2 seems to be reasonable, given that the parent cell should 
split in half when dividing. A favoring of values less than two is more than 
likely due to the growth of daughter cells following division but prior to the 
next image being taken (the image of the parent and the image of the daughter 
are four hours apart).

**2.** Check number of cells that remain to last time point.
```{r}
#| label: print-survivors

surviving_cells <- data.frame()
for(data in names(merged)) {
  df <- get(data)
  #Get list of original cells
  merged_0 <- filter(df, t == 0, trackId != -1)
  orig_cells <- merged_0$trackId
  l <- length(orig_cells)
  #Get list of cells at last time point
  t_max <- max(df$t)
  merged_max <- filter(df, t == t_max)
  #Find survivors
  temp <- merged_max %>% 
    subset(trackId %in% orig_cells)
  surviving_cells <- rbind(temp, surviving_cells)
  ls <- nrow(temp)
  print(paste0("Of ", l, " original cells in ", df[1,63], ", ", ls,
             " remained at the last time point."))
}
```

**3.** Check cell size over time for original cells.

```{r}
#| label: plot-survivors
#| fig-cap: Change in size of persisting cells over time. 
#| warning: false

for(data in names(merged)) {
  df <- get(data)
  survivors <- 
  survivors <- surviving_cells$trackId
  survivors <- df %>% 
  subset(trackId %in% survivors)
  survivors <- filter(survivors, trackId != -1)
  print(ggplot(survivors) + 
    geom_line(aes(t, Object_Area_0, color = as.character(trackId))) + 
    xlab("time") + 
    ylab("size in pixels") + 
    labs(color = "trackId") +
    ggtitle(paste0("Change in Size of ", df[1,63])) + 
    theme_light())
}
```
Persisting cells, despite their lack of division, have multiple, sudden changes
in size. This could be indicative of a missed mitotic cycle or a mislabeling 
onto another cell. 

```{r}
#| label: plot-size
#| fig-cap: Change in size of all cells over all time (left) and parent cells just before division (right).
#| warning: false

for(data in names(merged)) {
  df <- get(data)
  #Finding the difference in size across time
  sorted_merged <- df %>% 
    arrange(trackId) %>% 
    filter(trackId != -1)
  diff_merged <- sorted_merged %>% 
    group_by(trackId) %>% 
    arrange(t) %>% 
    mutate(diff = Object_Area_0 - lag(Object_Area_0, 
                                      default = first(Object_Area_0)))
  #Visualize
  a <- ggplot(diff_merged, aes(x = diff)) + 
    geom_histogram(bins = 100) + 
    theme_light() +
    geom_vline(aes(xintercept=mean(diff)), 
               linetype="dashed", 
               color = "blue", 
               size=1) + 
    xlab("change in size in pixels") + 
    ggtitle(paste0("All Cells of ", df[1,63]))
  #Find when each cell divides, then merge with previous table
  sorted_parentsmax <- parents_max %>% 
    arrange(trackId) %>% 
    select(trackId, div = t)
  assign(paste("sorted_parentsmax", data, sep = "_") , sorted_parentsmax)
  diff_div <- left_join(diff_merged, sorted_parentsmax)
  diff_div_1 <- filter(diff_div, t == div | t == (div-1)) %>% 
    arrange(trackId)
  #Visualize
  b <- ggplot(diff_div_1, aes(x = diff)) + 
    geom_histogram(bins = 100) + 
    geom_vline(aes(xintercept=mean(diff)), 
               linetype="dashed", 
               color = "blue", 
               size=1) +
    xlab("change in size in pixels") + 
    theme_light() + 
    ggtitle(paste0("Parents Before Division of ", df[1,63]))
  print(a+b)
}
```
My intuition would be that parents would lean towards either: (1) an increase 
in size prior to division, due to mitosis or general growth or (2) a decrease 
in size due to the cell pulling from the glass, lowering the surface area.

The overall change in size is dramatic in some cases. This is a phenotype that
is unlikely to change due to illumination differences or other image to image
variation. Instead, the extreme changes could be due to incorrect segmentation, 
unaccounted for cell fusion, or labels transferring between cells. 

**4.** Check changes in position

```{r}
#| label: plot-position
#| fig-cap: Visualizing change in position (x solid, y dashed) over time.
#| warning: false
#| message: false

for(data in names(merged)) {
  df <- get(data)
  #Finding the difference in size across time
  sorted_merged <- df %>% 
    arrange(trackId) %>% 
    filter(trackId != -1)
  #Find change in position
  diff_merged <- sorted_merged %>% 
    group_by(trackId) %>% 
    arrange(t) %>% 
    mutate(diff_x = x - lag(x, default = first(x)), 
           diff_y = y - lag(y, default = first(y)))
  assign(paste("diff", data, sep = "_") , diff_merged) 
  #Visualize
  a <- ggplot(diff_merged) + 
    geom_line(aes(t, diff_x, color = as.character(trackId)), linetype = 1) +
    geom_line(aes(t, diff_y, color = as.character(trackId)), linetype = 2) +
    theme_light() +
    theme(legend.position = "none") + 
    xlab("time") + 
    ylab("change in position in pixels") + 
    ggtitle(paste0("Position of All Cells of ", df[1,63]))
  b <- ggplot(diff_merged) + 
    geom_histogram(aes(x = diff_y), fill = "blue", alpha = 0.5, bins = 100) +
    geom_histogram(aes(x = diff_x), fill = "red", alpha = 0.5, bins = 100) + 
    theme_light() +
    xlim(-75,75) + 
    ggtitle(paste0("Position of All Cells of ", df[1,63])) + 
    xlab("change in position in pixels")
  print(a+b)
}
```
A sharp change in position greater than a cell can realistically move in that 
time could indicate a label hopping between cells.

```{r}
#| label: plot-parent-position
#| fig-cap: Visualizing change in position (x red,  y blue).
#| warning: false
#| message: false

for(data in names(merged)) {
  df <- get(data)
  diff_merged <- get(paste0("diff_", data))
  sorted_parentsmax <- get(paste0("sorted_parentsmax_", data))
  diff_div_par <- left_join(diff_merged, sorted_parentsmax)
  diff_div_par <- filter(diff_div_par, t == div | t == (div-1)) %>% 
    arrange(trackId)
  print(ggplot(diff_div_par) + 
    geom_histogram(aes(x = diff_y), fill = "blue", alpha = 0.5, bins = 100) +
    geom_histogram(aes(x = diff_x), fill = "red", alpha = 0.5, bins = 100) +
    xlim(-75,75) + 
    xlab("change in position in pixels") + 
    theme_light() + 
    ggtitle(paste0("Parents Before Division of ", df[1,63])))
}
```
A sharp change in position greater than a cell can realistically move in that 
time could indicate a label hopping between cells. The top images are Set 1,
and the bottom images are Set 2.

**5.** Check distances between cell centers of daughters after division

```{r}
#| label: plot-daughter-distance
#| fig-cap: The distance between the centers of daughter cells just following division.
#| message: false
#| warning: false

for(data in names(merged)) {
  df <- get(data)
  daughters_df <- get(paste0("daughters_", data))
  diff_daughters <- daughters_df %>% 
    group_by(parentTrackId) %>% 
    arrange(lineageId) %>% 
    mutate(diff_x = abs(x - lag(x, default = first(x))), 
           diff_y = abs(y - lag(y, default = first(y)))) %>% 
    filter(diff_x != 0)
  diff_daughters <- mutate(diff_daughters, 
                            distance = sqrt(((diff_x^2)+(diff_y^2))))
  print(ggplot(diff_daughters) + 
    geom_histogram(aes(distance), fill = "darkred", bins = 100) + 
    xlab("distance between centers in pixels") + 
    ggtitle(paste0("Distance Between Daughter Cells for ", df[1,63])) + 
    xlim(c(0,80)) +
    theme_light() +
    geom_vline(aes(xintercept=mean(distance)), linetype="dashed", size=1))
}
```
This analysis is similar to the one previous; the daughter cells should only be 
so far from one another after division. Cells that are too far apart may not 
actually be related.

**6.** Check phenotype of cells

```{r}
#| label: plot-phenotypes
#| fig-cap: Proportions of phenotypes at time just after division (daughters), time just before division (parents), and all times.

for(data in names(merged)) {
  df <- get(data)
  daughters_df <- get(paste0("daughters_", data))
  parents_max <- get(paste0("parentsmax_", data))
  a <- ggplot(daughters_df) + 
    geom_bar(aes(Classifier.Phenotype), fill = "darkred") + 
    theme_light() + 
    xlab("class") + 
    theme(axis.text.x = element_text(angle = 270)) +
    ggtitle(paste0("Daughters of ", df[1,63]))
  b <- ggplot(parents_max) + 
    geom_bar(aes(Classifier.Phenotype), fill = "darkblue") + 
    theme_light() + 
    xlab("class") + 
    theme(axis.title.y = element_blank(), axis.text.x = element_text(angle = 270)) + 
    ggtitle(paste0("Parents of ", df[1,63]))
  c <- ggplot(df) + 
    geom_bar(aes(Classifier.Phenotype)) + 
    theme_light() + 
    xlab("class") + 
    theme(axis.title.y = element_blank(), axis.text.x = element_text(angle = 270)) + 
    ggtitle(paste0("All Cells of ", df[1,63]))
  print(a+b+c)
}
```
It is important to see that parent cells are being appropriately identified as 
alive, as dead cells do not divide. It is possible for daughter cells to die 
following division, but it is concerning that the proportions of these match 
that of the parents and the general population.
```{r}
#| label: plots-phenotype
#| warning: false
#| message: false

#Create a "is not in" function
'%!in%' <- function(x,y)!('%in%'(x,y))

#Find cells at last time points, then label as parents/dividing.
for(data in names(merged)) {
  df <- get(data)
  #Find end point class
  df_1 <- filter(df, trackId != -1)
  cells_list <- unique(df_1$trackId)
  #Recreate parents list
  parents_list <- unique(df$parentTrackId)
  parents_list <- parents_list[!parents_list %in% 0]
  allcells_max <- data.frame()
  for(i in cells_list) {
    p <- filter(df, trackId == i)
    q <- max(p$t)
    r <- filter(p, t == q)
    if(i %in% parents_list) {
      r <- mutate(r, parent = TRUE)
      allcells_max <- rbind(allcells_max, r)
    }
    if(i %!in% parents_list) {
      r <- mutate(r, parent = FALSE)
      allcells_max <- rbind(allcells_max, r)    
    }
  }
  #Visualize class of cells at end points
  allcells_max <- filter(allcells_max, trackId != -1)
  print(ggplot(allcells_max) + 
    geom_bar(aes(Classifier.Phenotype, fill = parent), position = 'dodge') + 
    theme_light() + 
    xlab("class") + 
    ggtitle(paste0("Class of Cells at Last Time Point ", df[1,63])))
  #Get list of all time points labeled with dividing and parent statuses
  #Time consuming step
  allcells <- data.frame()
  for(i in cells_list) {
    p <- filter(df, trackId == i)
    q <- max(p$t)
    p <- cbind(p, lifetime = c(1:nrow(p)))
    r <- filter(p, t == q)
    s <- filter(p, t != q) %>% mutate(dividing = FALSE)
    if(i %in% parents_list) {
      r <- mutate(r, dividing = TRUE)
      r <- rbind(s, r)
      r <- mutate(r, parent = TRUE)
      allcells <- rbind(allcells, r)
    }
    if(i %!in% parents_list) {
      r <- mutate(r, dividing = FALSE)
      r <- rbind(s, r)
      r <- mutate(r, parent = FALSE)
      allcells <- rbind(allcells, r)    
    }
  }
  assign(paste("allcells", data, sep = "_") , allcells)
}

#Make graphs  
for(data in names(merged)) {
  df <- get(data)  
  #Visualize divisions
  merged_grouped <- df %>% 
    filter(trackId != -1) %>% 
    group_by(t) %>% 
    summarize(class = mean(Class))
  # ggplot(merged_grouped) + 
  #   geom_col(aes(t, class)) + 
  #   ylab("class") + 
  #   xlab("time") + 
  #   ggtitle("Average Phenotype over Time") + 
  #   theme_light()
  allcells <- get(paste0("allcells_", data))
  print(ggplot(filter(allcells, Class == 1)) + 
    geom_bar(aes(t, fill = dividing)) + 
    xlab("time") +
    ggtitle(paste0("Proportion of Live and Dividing Cells for ", df[1,63])) + 
    theme_light())
  print(ggplot(allcells) + 
    geom_bar(aes(t, fill = Classifier.Phenotype)) + 
    theme_light() + 
    xlab("time") + 
    labs(fill = "class") +
    ggtitle(paste0("Cell Phenotypes Over Time of ", df[1,63])))
  dividing_cells <- allcells %>% 
    filter(dividing == TRUE) 
  print(ggplot(dividing_cells) + 
    geom_bar(aes(t, fill = Classifier.Phenotype)) + 
    theme_light() + 
    xlab("time") + 
    labs(fill = "class") +
    ggtitle(paste0("Divisions over Time of ", df[1,63])))
  print(ggplot(allcells) + 
  geom_histogram(aes(lifetime)) + 
  theme_light() + 
  ggtitle(paste0("Lifetime of ", df[1,63])))
}
```
These figures all point towards some important, general trends. The class of 
cells at their last time point is meant to show whether a cell is no longer 
being tracked because it has divided (alive, true) or because it has died 
(dead, false). A live cell at it's last time point, should it not be a parent,
would have lost it's tracking due to an error. A dead cell that is a parent has
been incorrectly classified as dead. 

The average phenotype graphs should remain the same, as they are based on the 
classifier data and not the tracking.

**6.** Ground truth-ish
```{r}
#| label: plot-comparison
#| fig-cap: The number of cells counted by hand and for each tracking set. Black line = identity line; blue line = regression.
#| message: false
#| warning: false

a <- ggplot(filter(Book1, class == 'alive')) + 
  geom_bar(aes(slice)) + 
  theme_light() + xlab("t") + 
  ggtitle("Hand Annotation")

b <- ggplot(filter(allcells_merged1, Class == 1)) + 
  geom_bar(aes(t)) +  
  theme_light() + 
  ggtitle(paste0(allcells_merged1[1,63], " count"))

e <- ggplot(filter(allcells_merged2, Class == 1)) + 
  geom_bar(aes(t)) +  
  theme_light() + 
  ggtitle(paste0(allcells_merged2[1,63], " count"))

a+b+e

hand_ann <- Book1 %>% filter(class == "alive") %>% 
  group_by(slice) %>% 
  summarise(count = n())

b_ann <- allcells_merged1 %>% filter(Class == 1) %>% 
  mutate(slice = t+1) %>% 
  group_by(slice) %>% 
  summarise(count_1 = n())

e_ann <- allcells_merged2 %>% filter(Class == 1) %>% 
  mutate(slice = t+1) %>% 
  group_by(slice) %>% 
  summarise(count_2 = n())

abe <- left_join(hand_ann, b_ann)
abe <- left_join(abe, e_ann)
abe <- filter(abe, slice != 54)

test1 <- cor.test(abe$count, abe$count_1, method = "pearson")[[4]]
test2 <- cor.test(abe$count, abe$count_2, method = "pearson")[[4]] 
mae1 <- mae(abe$count, abe$count_1)
mae2 <- mae(abe$count, abe$count_2)

ggplot(abe) +
  geom_point(aes(count, count_1, color = slice)) +
  theme_light() +
  xlab("hand count") +
  ylab("count") +
  geom_smooth(aes(count, count_1), method = "lm", se = FALSE) +
  geom_abline(intercept = 0, slope = 1) +
  ggtitle(paste0("Number of live cells tracked for ", allcells_merged1[1,63])) +
  labs(caption = paste0("Pearson correlation = ", test1, "\nMean absolute error = ", mae1))
       
ggplot(abe) +
  geom_point(aes(count, count_2, color = slice)) +
  theme_light() +
  geom_smooth(aes(count, count_2), method = "lm", se = FALSE) +
  geom_abline(a = 0, b = 1) +
  xlab("hand count") +
  ylab("count") +
  ggtitle(paste0("Number of live cells tracked for ", allcells_merged2[1,63])) +
  labs(caption = paste0("Pearson correlation = ", test2, "\nMean absolute error = ", mae2))
```


```{r}
#| label: plot-divisions
#| fig-cap: The number of cell divisions counted by hand and for each tracking set.
#| message: false
#| warning: false

c <- ggplot(Book2) + 
  geom_bar(aes(slice)) + 
  scale_y_continuous(limits = c(0,60)) + 
  theme_light() + 
  xlab("t") + 
  ggtitle("Hand Divisions") + 
  scale_x_continuous(limits = c(0,25))

d <- ggplot(filter(allcells_merged1, dividing == TRUE)) + 
  geom_bar(aes(t)) + 
  scale_y_continuous(limits = c(0,60)) + 
  theme_light() + 
  ggtitle(paste0(allcells_merged1[1,63], " Divisions"))  + 
  scale_x_continuous(limits = c(0,25))

f <- ggplot(filter(allcells_merged2, dividing == TRUE)) + 
  geom_bar(aes(t)) + 
  scale_y_continuous(limits = c(0,60)) + 
  theme_light() + 
  ggtitle(paste0(allcells_merged2[1,63], " Divisions"))  + 
  scale_x_continuous(limits = c(0,25))

c+d+f

hand_ann <- Book2 %>% 
  group_by(slice) %>% 
  summarise(count = n())

d_ann <- allcells_merged1 %>%
  filter(dividing == TRUE) %>% 
  mutate(slice = t+1) %>% 
  group_by(slice) %>% 
  summarise(count_1 = n())

f_ann <- allcells_merged2 %>% 
  filter(dividing == TRUE) %>% 
  mutate(slice = t+1) %>% 
  group_by(slice) %>% 
  summarise(count_2 = n())

cdf <- left_join(hand_ann, d_ann)
cdf <- left_join(cdf, f_ann)

test1 <- cor.test(cdf$count, cdf$count_1, method = "pearson")[[4]]
test2 <- cor.test(cdf$count, cdf$count_2, method = "pearson")[[4]]
mae1 <- mae(cdf$count, cdf$count_1)
mae2 <- mae(cdf$count, cdf$count_2)

ggplot(cdf) +
  geom_point(aes(count, count_1, color = slice)) +
  theme_light() +
  geom_smooth(aes(count, count_1), method = "lm", se = FALSE) +
  geom_abline(a = 0, b = 1) +
  xlab("hand count") +
  ylab("count") +
  ggtitle(paste0("Divisions tracked for ", allcells_merged1[1,63])) +
  labs(caption = paste0("Pearson correlation = ", test1, "\nMean absolute error = ", mae1))

ggplot(cdf) +
  geom_point(aes(count, count_2, color = slice)) +
  geom_smooth(aes(count, count_2), method = "lm", se = FALSE) +
  geom_abline(a = 0, b = 1) +
  theme_light() +
  xlab("hand count") +
  ylab("count") +
  ggtitle(paste0("Divisions tracked for ", allcells_merged2[1,63])) +
  labs(caption = paste0("Pearson correlation = ", test2, "\nMean absolute error = ", mae2))
```

**7.** Feature mapping
```{r}
#| label: fig-17
#| fig-cap: UMAP of cell features, Set 1.

#Note: the UMAPs have been "broken" for the most recent sets -> WIP

#Set 1
# allcells1.data <- allcells1 %>% 
#   drop_na() %>% 
#   select(!c(1:6, 19:22, 39:42, 49:54, 56:69)) %>% 
#   mutate(ID = row_number())
# rownames(allcells1.data) <- NULL
# 
# allcells1.labels <- allcells1 %>% 
#   drop_na %>% 
#   select(!ID) %>% 
#   mutate(ID = row_number())
# rownames(allcells1.labels) <- NULL
# 
# umap_fit <- allcells1.data %>% 
#   select(where(is.numeric)) %>% 
#   column_to_rownames("ID") %>% 
#   scale() %>% 
#   umap()
# 
# umap_df <- umap_fit$layout %>% 
#   as.data.frame() %>% 
#   rename(UMAP1 = "V1", UMAP2 = "V2") %>% 
#   mutate(ID = row_number()) %>%
#   inner_join(allcells1.labels, by = "ID")
# 
# umap_df %>% 
#   ggplot(aes(x = UMAP1, y = UMAP2, color = Class)) + 
#   geom_point() + 
#   labs(x = "UMAP1", y = "UMAP2", subtitle = "UMAP plot")
```

```{r}
#| label: plot-PCA
#| fig-cap: PCA of cell features.

for(data in names(merged)) {
  df <- get(data)  
  allcells <- get(paste0("allcells_", data))
  allcells.data <- allcells %>% 
    drop_na() %>% 
    select(!c(t:mergerLabelId, starts_with("Center"), starts_with("Terminal"), 
              starts_with("Object_Center"), (ID:Class), dividing, parent, 
              Convex_Hull_Center_0, Convex_Hull_Center_1, 
              starts_with("Principal"), x, y, Size_in_pixels_0, 
              starts_with("Bounding_Box_Minimum"))) %>% 
    select(where(is.numeric)) %>% 
    scale()
  corr <- cor(allcells.data)
  print(ggcorrplot(corr, tl.cex = 5, title = paste0(df[1,63], " correlations")))
  pca <- princomp(corr)
  # print(summary(pca))
  fviz_eig(pca, addlabels = TRUE)
  print(fviz_pca_var(pca, col.var = "cos2",
              gradient.cols = c("navy", "orchid", "palegreen"),
              repel = TRUE, title = paste0(df[1,63], " PCA")))
}
```

**8.** Distances

```{r}
#| label: plot-mahalanobis
#| fig-cap: Mahalanobis distance by phenotype.
#| message: false
#| warning: false

for(data in names(merged)) {
  df <- get(data)  
  allcells <- get(paste0("allcells_", data))
  allcells.data <- allcells %>% 
    drop_na() %>% 
    select(!c(t:mergerLabelId, starts_with("Center"),
              starts_with("Terminal"), Number_of_Holes_0,
              starts_with("Object_Center"), (ID:Class), dividing, parent, 
              Convex_Hull_Center_0, Convex_Hull_Center_1, lifetime,
              starts_with("Principal"), x, y, Size_in_pixels_0)) %>% 
    mutate(ID = row_number())
  allcells.labels <- allcells %>% 
    drop_na %>% 
    select(!ID) %>% 
    mutate(ID = row_number())
  mal <- allcells.data
  mal$mal <- mahalanobis(mal, colMeans(mal), cov(mal))
  mal <- inner_join(mal, allcells.labels)
  print(ggplot(mal) + 
    geom_line(aes(t, mal, color = as.character(trackId))) + 
    theme_light() +
    theme(legend.position = "none") +
    ggtitle(paste0("Mahalanobis distance for ", df[1,63])))
}
```
```{r}
#| label: plot-euclidean
#| fig-cap: Euclidean distance by cell.
#| message: false
#| warning: false

for(data in names(merged)) {
  df <- get(data)  
  allcells <- get(paste0("allcells_", data))
  allcells.data <- allcells %>% 
    drop_na() %>% 
    select(!c(t, labelimageId, lineageId:mergerLabelId, starts_with("Center"),
              starts_with("Terminal"), Number_of_Holes_0,
              starts_with("Object_Center"), (ID:Class), dividing, parent, 
              Convex_Hull_Center_0, Convex_Hull_Center_1, lifetime,
              starts_with("Principal"), x, y, Size_in_pixels_0)) %>% 
    mutate(ID = row_number())
  allcells.labels <- allcells %>% 
    drop_na %>% 
    select(!ID) %>% 
    mutate(ID = row_number())
  #Most time consuming step; approx 30 minutes
  cols <- data.frame()
  data_1 <- data.frame()
  for(cell in unique(allcells.data$trackId)) {
    df_1 <- filter(allcells.data, trackId == cell)
    means <- df_1 %>% summarise_all(mean)
    cols <- data.frame(temp = c(1:nrow(df_1)))
    for(column in colnames(df_1)){
      if(column != "trackId") {
        col <- select(df_1, column)
        name <- colnames(col)
        value <- means[,column]
        col <- transmute(col, math = value - col[,1])
        colnames(col)[colnames(col) == "math"] = name
        cols <- cbind(cols, col)
      }
    }
    cols <- mutate(cols, trackId = cell)
    data_1 <- rbind(data_1, cols)
  }
  data_b <- data_1
  scaled.data <- scale(data_1[2:34])
  scaled.data <- cbind(select(data_1, temp, trackId), scaled.data)
  scaled.data$row_mean <- rowMeans(scaled.data[,c(3:35)], na.rm=TRUE)
  
  a <- ggplot(scaled.data) + 
    geom_line(aes(temp, row_mean, color = as.character(trackId))) +
    theme_light() +
    theme(legend.position = "none") +
    xlab("lifetime") +
    ylab("distance from mean") +
    ggtitle("Distance from mean over lifetime for ", df[1,63])
  
  scaled.data <- rename(scaled.data, lifetime = temp)
  scaled.data <- scaled.data %>% select(lifetime, trackId, row_mean)
  ts.data <- left_join(allcells, scaled.data, by = c("lifetime", "trackId"))
  assign(paste("ts.data", data, sep = "_") , ts.data) 
  b <- ggplot(ts.data) + 
    geom_line(aes(t, row_mean, color = as.character(trackId))) +
    theme_light() +
    theme(legend.position = "none") +
    xlab("time") +
    ylab("distance from mean") +
    ggtitle("Distance from mean over time for ", df[1,63])
  print(a+b)
}
```

**9.** Time series features

```{r}
#| label: plot-time-series
#| fig-cap: Time series features.
#| message: false
#| warning: false

for(data in names(merged)) {
  df <- get(data)  
  ts.data <- get(paste0("ts.data_", data))
  ts.filt <- ts.data %>% select(t, trackId, row_mean) %>% drop_na() %>% 
    group_by(trackId)
  ts.filt$t <- as_date(ts.filt$t)
  ts.xts <- ts.filt %>%
           group_by(trackId) %>%
           pivot_wider(names_from = "trackId", values_from = "row_mean", 
           values_fill = list(row_mean = 0))
  names <- colnames(ts.xts)
  ts_xts <- xts(ts.xts[-1], as.Date(ts.xts$t, "%d"))
  tsxts_feat <- tsfeatures(ts_xts)
  #Visualize
  a <- ggplot(tsxts_feat) +
    geom_histogram(aes(trend)) +
    theme_light() +
    ggtitle(paste0("Time series features of ", df[1,63]))
  b <- ggplot(tsxts_feat) +
    geom_histogram(aes(spike)) +
    theme_light()
  c <- ggplot(tsxts_feat) +
    geom_histogram(aes(linearity)) +
    theme_light()
  d <- ggplot(tsxts_feat) +
    geom_histogram(aes(curvature)) +
    theme_light()
  e <- ggplot(tsxts_feat) +
    geom_histogram(aes(e_acf1)) +
    theme_light()
  f <- ggplot(tsxts_feat) +
    geom_histogram(aes(e_acf10)) +
    theme_light()
  g <- ggplot(tsxts_feat) +
    geom_histogram(aes(entropy)) +
    theme_light()
  h <- ggplot(tsxts_feat) +
    geom_histogram(aes(x_acf1)) +
    theme_light() +
    ggtitle(paste0("Time series features "))
  i <- ggplot(tsxts_feat) +
    geom_histogram(aes(x_acf10)) +
    theme_light() +
    ggtitle(paste0("of ", df[1,63]))
  j <- ggplot(tsxts_feat) +
    geom_histogram(aes(diff1_acf1)) +
    theme_light()
  k <- ggplot(tsxts_feat) +
    geom_histogram(aes(diff1_acf10)) +
    theme_light()
  l <- ggplot(tsxts_feat) +
    geom_histogram(aes(diff2_acf1)) +
    theme_light()
  m <- ggplot(tsxts_feat) +
    geom_histogram(aes(diff2_acf10)) +
    theme_light()
  print((a+b+c+d)/(e+f+g))
  print((h+i+j)/(k+l+m))
  print(plot(tsxts_feat$entropy, tsxts_feat$spike, main = paste0("Time Series Features of ", df[1,63])))
  tsxts_feat.data <- tsxts_feat %>% 
    select(!c(frequency:seasonal_period)) %>% 
    mutate(ID = row_number()) %>% 
    drop_na()
  # UMAP
  # tsxts_feat.labels <- tsxts_feat %>% 
  #   mutate(ID = row_number()) %>% 
  #   drop_na
  # umap_fit_ts <- tsxts_feat.data %>% 
  #   select(where(is.numeric)) %>% 
  #   column_to_rownames("ID") %>% 
  #   scale() %>% 
  #   umap()
  # umap_df_ts <- umap_fit_ts$layout %>% 
  #   as.data.frame() %>% 
  #   rename(UMAP1 = "V1", UMAP2 = "V2") %>% 
  #   mutate(ID = row_number()) %>%
  #   inner_join(tsxts_feat.labels, by = "ID")
  # print(umap_df_ts %>% 
  #   ggplot(aes(x = UMAP1, y = UMAP2, color = ID)) + 
  #   geom_point() + 
  #   labs(x = "UMAP1", y = "UMAP2", subtitle = "UMAP plot"))
}
```
The trend is the overall movement of the time series stripped of its periodic 
behaviors. It is meant to trace the long term pattern of a time series, and it
can be thought of in terms of general increasing or decreasing. 

The spike measures the “spikiness” of a time series, and is computed as the 
variance of the leave-one-out variances of the remainder component. The 
remainder is what remains once the trend and periodic behaviors have been 
removed from the data. This would (ideally) be composed of random spikes.

Linearity and curvature of a time series calculated based on the coefficients 
of an orthogonal quadratic regression. This form of regression is less sensitive
to oscillation (which is often produced by periodic behaviors).

Features containing "acf":
The autocorrelation function of the series, the differenced series, and the 
twice-differenced series. The autocorrelation function finds the correlation 
between a time series and itself one time period apart (lag 1). A high 
correlation means that previous time points are more able to be used to predict 
later time points.

Entropy measures the “forecastability” of a time series, where low values 
indicate a high signal-to-noise ratio, and large values occur when a series is 
difficult to forecast.

## Tracking visualization

**1.** Showing all cells over time, colored by days alive. 

```{r}
# #Sets 1 & 2
# palette(rainbow(53))
# fi <- list.files(paste0(root, filesep, "B01_20230407_Incucyte_Images_woGFP_Analysis_QI_Core"),pattern=well,full.names = T)
# fi <- fi[2:54]
# merged=list(merged1=merged1, merged2=merged2)
# for(set in names(merged)){
#   for(i in unique(allcells2$t)){
#     p <- allcells2 %>% filter(t == i, trackId != -1)
#     tiff(paste0("~/Downloads/track", merged[[set]][1,"ID"], "_", i, ".tif"), width = 1408, height = 1040)
#     img=bioimagetools::readTIF(fi[i+1],as.is = T)
#     img <- EBImage::resize(img, dim(img)[1]/1)
#     plot(raster::as.raster(img[,,,1]))
#     points(p$x, p$y, pch=(p$Class+2), col = p$lifetime, cex=1.5, lwd = 2)
#     text(p$x+20, p$y, labels=p$trackId, cex=1)
#     mtext(fileparts(fi[i+1])$name)
#     legend("topleft", legend = unique(allcells2$Classifier.Phenotype), pch=unique(allcells2$Class+2), cex = 1)
#     dev.off()
#   }
# }

```
**2.** Show one track at a time.
```{r}
# ## visualize one track at a time on raw images
# # fi=list.files(paste0(root, filesep, "B01_20230407_Incucyte_Images_woGFP_Analysis_QI_Core"),pattern=well,full.names = T)
# # fi=list.files(paste0(root, filesep, "I01_20231219_HALOCellClassification/E2_1_Registered_image"),pattern=well,full.names = T)
# fi=list.files(paste0(root, filesep, "I01_20231219_HALOCellClassification/E2_1_Overlays"),pattern=well,full.names = T)
# # trackIDs=names(la)[order(sapply(la,nrow),decreasing = T)]
# trackIDs=as.character(order(tsxts_feat$linearity,decreasing = T))
# trackIDs=trackIDs[trackIDs %in% names(la1)]
# # goodTracks = c("136")
# # match(goodTracks,trackIDs)
# # badTracks = c("1057")
# # match(badTracks,trackIDs)
# for(trackID in c(head(trackIDs),tail(trackIDs))){
#   toi=la1[[trackID]]
#   pdf(paste0("~/Downloads/testTrack_",trackID,".pdf"), width=14, height = 9)
#   
#   plot(toi$t,toi$Object_Area_0, col=toi$Class+2,pch=20); 
#   legend("topleft",unique(toi$Classifier.Phenotype), fill=unique(toi$Class+2))
#   
#   fidx=max(2,grep(toi$time[1],fi,value=F))
#   center=NULL
#   # for(i in 1:nrow(toi)){
#   while(fidx<=length(fi)){  
#     thistime=strsplit(fileparts(fi[fidx])$name,"_")[[1]][4]
#     i=which(toi$time==thistime)
#     if(is.null(center)){
#       center=c(toi$x[i], toi$y[i])
#     }
#     img=bioimagetools::readTIF(fi[fidx],as.is = T)
#     img <- EBImage::resize(img, dim(img)[1]/1)
#     par(mfrow=c(1,2))
#     plot(raster::as.raster(img[,,,1]), xlim=c(center[1]-200,center[1]+200), ylim=c(center[2]-50,center[2]+50))
#     mtext(fileparts(fi[fidx])$name)
#     ## Just for testing
#     # tmp=merged[merged$t==toi$t[i],]
#     # points(tmp$x,tmp$y,pch=3,col="green")
#     if(!isempty(i)){
#       points(toi$x[i], toi$y[i],pch=3,cex=3)
#     }
#     plot(toi$t,toi$Object_Area_0, col=c("black","red")[1+(toi$time==thistime)],pch=20,cex=2); 
#     fidx=fidx+1;
#   }
#   dev.off()
# }
```
**3.** Show overlays and tracking ID.
```{r}
# ## plot overlays along with tracking ID:
# pdf(paste0("~/Downloads/trackIDs.pdf"))
# fidx=2
# while(fidx<=length(fi)){  
#   img=bioimagetools::readTIF(fi[fidx],as.is = T)
#   img <- EBImage::resize(img, dim(img)[1]/1)
#   plot(raster::as.raster(img[,,,1]), xlim=c(211,700), ylim=c(150,650))
#   mtext(fileparts(fi[fidx])$name)
#   
#   thistime=strsplit(fileparts(fi[fidx])$name,"_")[[1]][4]
#   tmp=sapply(names(la1), function(x) la1[[x]][la1[[x]]$time==thistime,,drop=F], simplify = F)
#   tmp=sapply(names(tmp), function(x) cbind(tmp[[x]], rep(x, nrow(tmp[[x]]))), simplify = F)
#   tmp=do.call(rbind, tmp)
#   tmp=as.data.frame(tmp)
#   tmp=tmp[which(!sapply(tmp$x,isempty)),]
#   colnames(tmp)[ncol(tmp)]="lineageID"
#   text(unlist(tmp$x), unlist(tmp$y),labels = unlist(tmp$lineageID), cex=0.6)
#   fidx=fidx+1;
# }
# dev.off()
```