---
title: "Tracking QC Analysis"
author: "Jordan Albrecht, Noemi Andor"
format: 
  html:
    self-contained: true
    embed-resources: true
    code-fold: true
    code-tools: true
    df-print: paged
editor: source
theme: sandstone
whitespace: large
---

## Data

**Plate map**: An xlsx with variables Row, Ploidy, and SUM159 of varying concentrations of treatment. This study has four replicates of 2N and 4N SUM159.

```{r}
#| echo: false
#| warning: false
library(readxl)
library(matlab)
library(lubridate)
root="~/Projects/PMO/HighPloidy_DoubleEdgedSword/data/BreastCancerCLs/SUM159/K00_GemcitabineExposure_033023"
path2repo="~/Repositories/Gemcitabine-model"
Gemcitabine_PlateMap_20230327 <- read_excel(paste0(root, filesep, "Gemcitabine_PlateMap_20230327.xlsx"))
Gemcitabine_PlateMap_20230327

```

**Ilastik Output**: A csv with IDs etc

**Classifier Output**: 

## Tracking Results Analysis

Call the Ilastik tracking file.

```{r}
#| warning: false
#| message: false

#Requisite loading of packages
library(tsfeatures)
library(matlab)
library(xts)
library(celltrackR)
library(tidyverse)
library(patchwork)
library(umap)
library(corrr)
library(ggcorrplot)
library(FactoMineR)
library(factoextra)


#Set to folder containing images, tracking results, plate table...
setwd(paste0(root, filesep, "/H02_20230831_CellTracking_Ilastik"))
#Import hand annotations
Book1 <- read_excel(paste0(path2repo, filesep, "Data/Book1.xlsx"))
Book2 <- read_excel(paste0(path2repo, filesep, "Data/Book2.xlsx"))

#Retired file pathways
#dat1 <- read.csv(paste0(root, filesep, "/H02_20230831_CellTracking_Ilastik/E2_1_T=50_App=1001_Disapp=1.5_CSV-Table.tiff.csv"))
#dat2 <- read.csv(paste0(root, filesep, "/H02_20230831_CellTracking_Ilastik/E2_1_T=50_App=1001_Disapp=1001_CSV-Table.tiff.csv"))
#dat1 <- mutate(dat1, ID = "E2_1_T=50_App=1001_Disapp=1.5")
#dat2 <- mutate(dat2, ID = "E2_1_T=50_App=1001_Disapp=1001")

#Open the two tracking result tables for comparison
dat1 <- read.csv(paste0(root, filesep, "/Tracking_12-15/E2_1_CSV-Table_P1.h5.csv"))
dat2 <- read.csv(paste0(root, filesep, "/Tracking_12-15/E2_1_CSV-Table_P2.h5.csv"))
dat1 <- mutate(dat1, ID = "E2_1_P2")
dat2 <- mutate(dat2, ID = "E2_1_P3")

#Read in classifier file
# seg=read.csv(paste0(root, filesep, "/B01_20230407_Incucyte_Images_woGFP_Analysis_QI_Core/Results_Classifier_2/E_objectdata.csv"))
seg=read.csv(paste0(root, filesep, "/I01_20231219_HALOCellClassification/E2_1_Total_Object_Results.csv"))

#Set the name of the well
well= "E2_1_"
```

Create the tracking plot.

```{r}
#| warning: false
#| message: false
#| label: fig-1
#| fig-cap: Trajectories of Set 1.

#Get the dimensions of the frame of motion for every cell?
ii=match(c("frame","Bounding_Box_Maximum_0","Bounding_Box_Maximum_1"),colnames(dat1))
jj=match(c("frame","Bounding_Box_Maximum_0","Bounding_Box_Maximum_1"),colnames(dat2))

colnames(dat1)[ii]=c("t","x","y")
colnames(dat2)[jj]=c("t","x","y")
#List dfs of cells in well
la=sapply(unique(dat1$trackId), function(x) dat1[dat1$trackId==x,c("t","x","y","Object_Area_0")], simplify = F)
la2=sapply(unique(dat2$trackId), function(x) dat2[dat2$trackId==x,c("t","x","y","Object_Area_0")], simplify = F)
#Calls cells as their trackID
names(la)=as.character(unique(dat1$trackId))
names(la2)=as.character(unique(dat2$trackId))
la=la[!names(la) %in% c("-1")]
la2=la2[!names(la2) %in% c("-1")]
plot(as.tracks(la))
```
```{r}
#| warning: false
#| message: false
#| label: fig-2
#| fig-cap: Trajectories of Set 2.

plot(as.tracks(la2))
```

## Merge with classifier results

**1.** Read in classifier data

```{r}
#Pull in the data for the well of interest
seg=seg[grep(well,seg$Image.Location),]
seg$time=sapply(strsplit(seg$Image.Location,"_"), function(x) x[length(x)])
seg$time=gsub(".tif","",seg$time)
```

**2.** Change times to frames

```{r}
#Replace time with frame
map=0:(length(unique(seg$time))-1)
names(map)=sort(unique(seg$time))
seg$t=map[seg$time]
seg=seg[order(seg$t),]

#Check changes
boxplot(seg$t~seg$time, horizontal = T,las=2,ylab="",cex.axis=0.54)
```

**3.** Compare x and y values

```{r}
#Pull x and y columns
ii=match(c("XMax","YMax"),colnames(seg))
colnames(seg)[ii]=c("x","y")
coi1=intersect(colnames(seg),colnames(dat1))
coi2=intersect(colnames(seg),colnames(dat2))
ii=sapply(map, function(t) list(seg=which(seg$t==t), dat1=which(dat1$t==t)), simplify = F)
jj=sapply(map, function(t) list(seg=which(seg$t==t), dat2=which(dat2$t==t)), simplify = F)
ii=ii[sapply(ii, function(x) min(length(x[[1]]),length(x[[2]])))>0]
jj=jj[sapply(jj, function(x) min(length(x[[1]]),length(x[[2]])))>0]
d1=sapply(ii, function(t) flexclust::dist2(seg[t$seg,coi1,drop=F],dat1[t$dat1,coi1,drop=F]))
d2=sapply(jj, function(t) flexclust::dist2(seg[t$seg,coi2,drop=F],dat2[t$dat2,coi2,drop=F]))
```

**4.** Assign tracked cells to segmented cells.

```{r}
#Combining dfs using names and x values ?
seg_matched=sapply(names(d1), function(x) seg[ii[[x]]$seg,][apply(d1[[x]],2,which.min),],simplify = F)
merged1=sapply(names(d1), function(x) cbind(dat1[ii[[x]]$dat1,],  seg_matched[[x]][,c("time","Classifier.Phenotype")]), simplify = F)
merged_B1=merged1
merged1=do.call(rbind, merged1)
merged1=merged1[order(merged1$t),];

seg_matched=sapply(names(d2), function(x) seg[jj[[x]]$seg,][apply(d2[[x]],2,which.min),],simplify = F)
merged2=sapply(names(d2), function(x) cbind(dat2[jj[[x]]$dat2,], seg_matched[[x]][,c("time","Classifier.Phenotype")]), simplify = F)
merged_B2=merged2
merged2=do.call(rbind, merged2)
merged2=merged2[order(merged2$t),];
```

**5.** Find the center of each cell.

```{r}
merged1$y=(merged1$Bounding_Box_Minimum_1+merged1$y)/2
merged1$x=(merged1$Bounding_Box_Minimum_0+merged1$x)/2
merged1$y=-merged1$y
merged1$y=merged1$y-min(merged1$y)
merged2$y=(merged2$Bounding_Box_Minimum_1+merged2$y)/2
merged2$x=(merged2$Bounding_Box_Minimum_0+merged2$x)/2
merged2$y=-merged2$y
merged2$y=merged2$y-min(merged2$y)

#View changes
select(merged1, labelimageId, mergerLabelId, Object_Center_0, Object_Center_1, x, y, Bounding_Box_Minimum_0, Bounding_Box_Minimum_1, Center_of_the_object_0, Center_of_the_object_1)
select(merged2, labelimageId, mergerLabelId, Object_Center_0, Object_Center_1, x, y, Bounding_Box_Minimum_0, Bounding_Box_Minimum_1, Center_of_the_object_0, Center_of_the_object_1)
```

The 'x' and 'y' values in the merged table were brought in from the classifier, and the 'Bounding_Box' values are from the tracking set ?

**6.** Make Dead, Transitional, and Alive statuses into a range from -1 to 1.

```{r}
merged1$Class=1
merged1$Class[merged1$Classifier.Phenotype=="Transitional"]=0
merged1$Class[merged1$Classifier.Phenotype=="Dead"]=-1
merged2$Class=1
merged2$Class[merged2$Classifier.Phenotype=="Transitional"]=0
merged2$Class[merged2$Classifier.Phenotype=="Dead"]=-1
```

## Existing Quality Control

**1.** See if number of tracked cells is same as number of segmented cells.

```{r}
#| label: fig-3
#| fig-cap: Pearson correlation for E2_1_T=50_App=1001_Disapp=1.5
#| warning: false

#Set 1
sapply(ii, function(x) c(length(x$dat1),length(x$seg)))

#Pull tracked cells
la1=sapply(unique(merged1$trackId), function(x) merged1[merged1$trackId==x,c("t","x","y","Object_Area_0","time","Classifier.Phenotype","Class")], simplify = F)
names(la1)=as.character(unique(merged1$trackId))
la1=la1[!names(la1) %in% c("-1")]
la_1=la1[sapply(la1,nrow)>=10]

#Uhhh
# la_1=sapply(la_1, function(x) x[x$t<(24*10)/4,],simplify = F)
# la_1=la_1[sapply(la_1,nrow)>=10 & sapply(la_1, function(x) min(x$t))<100]
te1=sapply(la_1, function(x) cor.test(x$t,x$Class)$estimate)
hist(te1, xlab="Pearson correlation(time, cell state)",main="Correlation between time and assigned class")
mtext("Dead cells should not come back to life (corr should be negative for dying cells)")
mtext("corr should be near-zero for cells that never die",line = 3)
la_1=la_1[is.na(te1) | te1<0.05];
la_1=la_1[!sapply(la_1,is.null)];
print(paste(length(la_1)/length(te1),"cells included (which do not come back to live from dead)!"))
```

```{r}
#| label: fig-4
#| fig-cap: Pearson correlation for E2_1_T=50_App=1001_Disapp=1001
#| warning: false

#Set 2
sapply(jj, function(x) c(length(x$dat2),length(x$seg)))

#Pull tracked cells
la2=sapply(unique(merged2$trackId), function(x) merged2[merged2$trackId==x,c("t","x","y","Object_Area_0","time","Classifier.Phenotype","Class")], simplify = F)
names(la2)=as.character(unique(merged2$trackId))
la2=la2[!names(la2) %in% c("-1")]

la_2=sapply(la2, function(x) x[x$t<(24*10)/4,],simplify = F)
la_2=la_2[sapply(la_2,nrow)>=10 & sapply(la_2, function(x) min(x$t))<100]
te2=sapply(la_2, function(x) cor.test(x$t,x$Class)$estimate)
hist(te2, xlab="Pearson correlation(time, cell state)",main="Correlation between time and assigned class")
mtext("Dead cells should not come back to life (corr should be negative for dying cells)")
mtext("corr should be near-zero for cells that never die",line = 3)
la_2=la_2[is.na(te2) | te2<0.05];
la_2=la_2[!sapply(la_2,is.null)];
print(paste(length(la_2)/length(te2),"cells included (which do not come back to live from dead)!"))
```

**2.** Visualize change in size over time. Removed due to disturbingly long null output.

**3.** Visualize one track at a time on raw images. This cannot be run due to missing files.

## New Quality Control

**1.** Check size of parent and daughters for cell division events.
```{r}
#| warning: false
#| label: fig-5
#| fig-cap: Ratio of parent cell size to average daughter size for Set 1

#Set 1
#Find a list of all the parents
parents_list1 <- unique(merged1$parentTrackId)
parents_list1 <- parents_list1[!parents_list1 %in% 0]

#Find all of the data points for each parent
parents_df1 <- merged1 %>% subset(trackId %in% parents_list1)
#Find all of the daughters of each parent in the list
daughters_df1 <- merged1 %>% subset(parentTrackId %in% parents_list1)

#Finding only the areas of parents at latest time point (right before division)
parents_max1 <- data.frame()
for(i in parents_list1){
  p <- filter(parents_df1, trackId == i)
  q <- max(p$t)
  r <- filter(p, t == q)
  parents_max1 <- rbind(parents_max1, r)
}

#Finding the average area of each daughter
daughters_means1 <- data.frame()
for(i in parents_list1){
  p <- filter(daughters_df1, parentTrackId == i)
  daughtersArea1 <- mean(p$Object_Area_0)
  r <- cbind(trackId = i, daughtersArea1)
  daughters_means1 <- rbind(daughters_means1, r)
}

#Visualize
ratios1 <- left_join(parents_max1, daughters_means1) 
ratios1 <- ratios1 %>% 
  mutate(ratio = Object_Area_0/daughtersArea1)

ggplot(ratios1, aes(x = ratio)) + 
  geom_histogram(bins = 150) + 
  xlim(0, 6) +
  geom_vline(aes(xintercept=mean(ratio)), 
             color="blue", 
             linetype="dashed", 
             size=1) + 
  theme_light() + 
  ggtitle("Ratio of Parent Cell Size to Average Daughter Cell Size")
```
```{r}
#| warning: false
#| label: fig-6
#| fig-cap: Ratio of parent cell size to average daughter size for Set 2.

#Set 2
#Find a list of all the parents
parents_list2 <- unique(merged2$parentTrackId)
parents_list2 <- parents_list2[!parents_list2 %in% 0]

#Find all of the data points for each parent
parents_df2 <- merged2 %>% subset(trackId %in% parents_list2)
#Find all of the daughters of each parent in the list
daughters_df2 <- merged2 %>% subset(parentTrackId %in% parents_list2)

#Finding only the areas of parents at latest time point (right before division)
parents_max2 <- data.frame()
for(i in parents_list2){
  p <- filter(parents_df2, trackId == i)
  q <- max(p$t)
  r <- filter(p, t == q)
  parents_max2 <- rbind(parents_max2, r)
}

#Finding the average area of each daughter
daughters_means2 <- data.frame()
for(i in parents_list2){
  p <- filter(daughters_df2, parentTrackId == i)
  daughtersArea2 <- mean(p$Object_Area_0)
  r <- cbind(trackId = i, daughtersArea2)
  daughters_means2 <- rbind(daughters_means2, r)
}

#Visualize
ratios2 <- left_join(parents_max2, daughters_means2) 
ratios2 <- ratios2 %>% 
  mutate(ratio = Object_Area_0/daughtersArea2)

ggplot(ratios2, aes(x = ratio)) + 
  geom_histogram(bins = 150) + 
  xlim(0, 6) +
  geom_vline(aes(xintercept=mean(ratio)), 
             color="blue", 
             linetype="dashed", 
             size=1) + 
  theme_light() + 
  ggtitle("Ratio of Parent Cell Size to Average Daughter Cell Size")
```
An average ratio of 2 seems to be reasonable, given that the parent cell should 
split in half when dividing. A favoring of values less than two is more than 
likely due to the growth of daughter cells following division but prior to the 
next image being taken (the image of the parent and the image of the daughter 
are four hours apart).

**2.** Check number of cells that remain to last time point.
```{r}
#Set 1
#Get list of original cells
merged_01 <- filter(merged1, t == 0, trackId != -1)
orig_cells1 <- merged_01$trackId
l1 <- length(orig_cells1)

#Get list of cells at last time point
t_max <- max(merged1$t)
merged1_max <- filter(merged1, t == t_max)
#Find survivors
surviving_cells1 <- merged1_max %>% 
  subset(trackId %in% orig_cells1)
ls1 <- nrow(surviving_cells1)

print(paste0("Of ", l1, " original cells in ", merged1[1,63], ", ", ls1,
             " remained at the last time point."))

#Set2
#Get list of original cells
merged_02 <- filter(merged2, t == 0, trackId != -1)
orig_cells2 <- merged_02$trackId
l2 <- length(orig_cells2)

#Get list of cells at last time point
t_max <- max(merged2$t)
merged_max2 <- filter(merged2, t == t_max)
#Find survivors
surviving_cells2 <- merged_max2 %>% 
  subset(trackId %in% orig_cells2)
ls2 <- nrow(surviving_cells2)

print(paste0("Of ", l2, " original cells in ", merged2[1,63], ", ", ls2,
             " remained at the last time point."))
```

**3.** Check cell size over time for original cells.

```{r}
#| label: fig-6.b
#| fig-cap: Change in size of persisting cells over time. 
#| warning: false
survivors1 <- surviving_cells1$trackId
survivors1 <- merged1 %>% 
  subset(trackId %in% survivors1)
survivors1 <- filter(survivors1, trackId != -1)

ggplot(survivors1) + 
  geom_line(aes(t, Object_Area_0, color = as.character(trackId))) + 
  xlab("time") + 
  ylab("size in pixels") + 
  labs(color = "trackId") +
  ggtitle("Change in Size Set 1") + 
  theme_light()

survivors2 <- surviving_cells2$trackId
survivors2 <- merged2 %>% 
  subset(trackId %in% survivors2)
survivors2 <- filter(survivors2, trackId != -1)

ggplot(survivors2) + 
  geom_line(aes(t, Object_Area_0, color = as.character(trackId))) + 
  xlab("time") + 
  ylab("size in pixels") + 
  ggtitle("Change in Size Set 2") + 
  labs(color = "trackId") +
  theme_light()
```
Persisting cells, despite their lack of division, have multiple, sudden changes
in size. This could be indicative of a missed mitotic cycle or a mislabeling 
onto another cell. 

```{r}
#| warning: false
#| label: fig-7
#| fig-cap: Change in size of all cells over all time (left) and parent cells just before division (right) for Set 1

#Set 1
#Finding the difference in size across time
sorted_merged1 <- merged1 %>% 
  arrange(trackId) %>% 
  filter(trackId != -1)

diff_merged1 <- sorted_merged1 %>% 
  group_by(trackId) %>% 
  arrange(t) %>% 
  mutate(diff = Object_Area_0 - lag(Object_Area_0, 
                                    default = first(Object_Area_0)))

#Visualize
a <- ggplot(diff_merged1, aes(x = diff)) + 
  geom_histogram(bins = 100) + 
  xlim(-600,600) + 
  theme_light() +
  geom_vline(aes(xintercept=mean(diff)), 
             linetype="dashed", 
             color = "blue", 
             size=1) + 
  xlab("change in size in pixels") + 
  ggtitle("All Cells")

#Find when each cell divides, then merge with previous table
sorted_parentsmax1 <- parents_max1 %>% 
  arrange(trackId) %>% 
  select(trackId, div = t)

diff_div1 <- left_join(diff_merged1, sorted_parentsmax1)
diff_div2_1 <- filter(diff_div1, t == div | t == (div-1)) %>% 
  arrange(trackId)

b <- ggplot(diff_div2_1, aes(x = diff)) + 
  geom_histogram(bins = 100) + 
  xlim(-600,600) +
  geom_vline(aes(xintercept=mean(diff)), 
             linetype="dashed", 
             color = "blue", 
             size=1) +
  xlab("change in size in pixels") + 
  theme_light() + 
  ggtitle("Parents Before Division")

a+b
```
My intuition would be that parents would lean towards either: (1) an increase 
in size prior to division, due to mitosis or general growth or (2) a decrease 
in size due to the cell pulling from the glass, lowering the surface area.

```{r}
#| warning: false
#| label: fig-8
#| fig-cap: Change in size of all cells over all time (left) and parent cells just before division (right) for Set 2. 

#Set 2
#Finding the difference in size across time
sorted_merged2 <- merged2 %>% 
  arrange(trackId) %>% 
  filter(trackId != -1)
diff_merged2 <- sorted_merged2 %>% 
  group_by(trackId) %>% 
  arrange(t) %>% 
  mutate(diff = Object_Area_0 - lag(Object_Area_0, default = 
                                      first(Object_Area_0)))

a <- ggplot(diff_merged2, aes(x = diff)) + 
  geom_histogram(bins = 100) + 
  xlim(-600,600) + theme_light() +
  geom_vline(aes(xintercept=mean(diff)), 
             linetype="dashed", 
             color = "blue", size=1) + 
  xlab("change in size in pixels") + 
  ggtitle("All Cells")

#Find when each cell divides, then merge with previous table
sorted_parentsmax2 <- parents_max2 %>% 
  arrange(trackId) %>% 
  select(trackId, div = t)

diff_div2 <- left_join(diff_merged2, sorted_parentsmax2)
diff_div2_2 <- filter(diff_div2, t == div | t == (div-1)) %>% arrange(trackId)

b <- ggplot(diff_div2_2, aes(x = diff)) + 
  geom_histogram(bins = 100) + 
  xlim(-600,600) +
  geom_vline(aes(xintercept=mean(diff)), 
             linetype="dashed", 
             color = "blue", 
             size = 1) +
  xlab("change in size in pixels") + 
  theme_light() + 
  ggtitle("Parents Before Division")

a+b
```
The overall change in size is dramatic in some cases. This is a phenotype that
is unlikely to change due to illumination differences or other image to image
variation. Instead, the extreme changes could be due to incorrect segmentation, 
unaccounted for cell fusion, or labels transferring between cells. 

**4.** Check changes in position

```{r}
#| warning: false
#| message: false
#| label: fig-9
#| fig-cap: Visualizing change in position (x solid, y dashed) over time.

#Set 1
#Find change in position
diff_merged1 <- sorted_merged1 %>% 
  group_by(trackId) %>% 
  arrange(t) %>% 
  mutate(diff_x = x - lag(x, default = first(x)), 
         diff_y = y - lag(y, default = first(y)))

a <- ggplot(diff_merged1) + 
  geom_line(aes(t, diff_x, color = as.character(trackId)), linetype = 1) +
  geom_line(aes(t, diff_y, color = as.character(trackId)), linetype = 2) +
  theme_light() +
  theme(legend.position = "none") + 
  xlab("time") + 
  ylab("change in position in pixels") + 
  ggtitle("Change in Position of All Cells Set 1")

b <- ggplot(diff_merged1) + 
  geom_histogram(aes(x = diff_y), fill = "blue", alpha = 0.5, bins = 100) +
  geom_histogram(aes(x = diff_x), fill = "red", alpha = 0.5, bins = 100) + 
  theme_light() +
  xlim(-75,75) + 
  ggtitle("Change in Position of All Cells") + 
  xlab("change in position in pixels")

#Set 2
#Find change in position
diff_merged2 <- sorted_merged2 %>% 
  group_by(trackId) %>% 
  arrange(t) %>% 
  mutate(diff_x = x - lag(x, default = first(x)), 
         diff_y = y - lag(y, default = first(y)))

c <- ggplot(diff_merged2) + 
  geom_line(aes(t, diff_x, color = as.character(trackId)), linetype = 1) +
  geom_line(aes(t, diff_y, color = as.character(trackId)), linetype = 2) +
  theme_light() +
  theme(legend.position = "none") + 
  xlab("time") + 
  ylab("change in position in pixels") + 
  ggtitle("Set 2")

d <- ggplot(diff_merged2) + 
  geom_histogram(aes(x = diff_y), fill = "blue", alpha = 0.5, bins = 100) +
  geom_histogram(aes(x = diff_x), fill = "red", alpha = 0.5, bins = 100) + 
  theme_light() +
  xlim(-75,75) + 
  ggtitle("Change in Position of All Cells") + 
  xlab("change in position in pixels")

a+c
```
A sharp change in position greater than a cell can realistically move in that 
time could indicate a label hopping between cells. The left figure is 
Set 1 and right is Set 2.

```{r}
#| warning: false
#| message: false
#| label: fig-10
#| fig-cap: Visualizing change in position (x red,  y blue).

#Set 1
diff_div_par1 <- left_join(diff_merged1, sorted_parentsmax1)
diff_div_par1 <- filter(diff_div_par1, t == div | t == (div-1)) %>% 
  arrange(trackId)

e <- ggplot(diff_div_par1) + 
  geom_histogram(aes(x = diff_y), fill = "blue", alpha = 0.5, bins = 100) +
  geom_histogram(aes(x = diff_x), fill = "red", alpha = 0.5, bins = 100) +
  xlim(-75,75) + 
  xlab("change in position in pixels") + 
  theme_light() + 
  ggtitle("Change of Parents Before Division")



#Set 2
diff_div_par2 <- left_join(diff_merged2, sorted_parentsmax2)
diff_div_par2 <- filter(diff_div_par2, t == div | t == (div-1)) %>% 
  arrange(trackId)

f <- ggplot(diff_div_par2) + 
  geom_histogram(aes(x = diff_y), fill = "blue", alpha = 0.5, bins = 100) +
  geom_histogram(aes(x = diff_x), fill = "red", alpha = 0.5, bins = 100) +
  xlim(-75,75) + 
  xlab("change in position in pixels") + 
  theme_light() + 
  ggtitle("Change of Parents Before Division")

b+e

d+f
```
A sharp change in position greater than a cell can realistically move in that 
time could indicate a label hopping between cells. The top images are Set 1,
and the bottom images are Set 2.

**5.** Check distances between cell centers of daughters after division

```{r}
#| label: fig-11
#| fig-cap: The distance between the centers of daughter cells just following division for Set 1 (left) and Set 2 (right).
#| message: false
#| warning: false

#Set 1
diff_daughters1 <- daughters_df1 %>% 
  group_by(parentTrackId) %>% 
  arrange(lineageId) %>% 
  mutate(diff_x = abs(x - lag(x, default = first(x))), 
         diff_y = abs(y - lag(y, default = first(y)))) %>% 
  filter(diff_x != 0)

diff_daughters1 <- mutate(diff_daughters1, 
                          distance = sqrt(((diff_x^2)+(diff_y^2))))

a <- ggplot(diff_daughters1) + 
  geom_histogram(aes(distance), fill = "darkred", bins = 100) + 
  xlab("distance between centers in pixels") + 
  ggtitle("Distance Between Daughter Cells") + 
  xlim(c(0,80)) +
  theme_light() +
  geom_vline(aes(xintercept=mean(distance)), linetype="dashed", size=1)


#Set 2
diff_daughters2 <- daughters_df2 %>% 
  group_by(parentTrackId) %>% 
  arrange(lineageId) %>% 
  mutate(diff_x = abs(x - lag(x, default = first(x))), 
         diff_y = abs(y - lag(y, default = first(y)))) %>% 
  filter(diff_x != 0)

diff_daughters2 <- mutate(diff_daughters2, 
                          distance = sqrt(((diff_x^2)+(diff_y^2))))

b <- ggplot(diff_daughters2) + 
  geom_histogram(aes(distance), fill = "darkred", bins = 100) +
  xlab("distance between centers in pixels") + 
  ggtitle("Distance Between Daughter Cells") + 
  xlim(c(0,80)) +
  theme_light() +
  geom_vline(aes(xintercept=mean(distance)), linetype="dashed", size=1)

a+b
```
This analysis is similar to the one previous; the daughter cells should only be 
so far from one another after division. Cells that are too far apart may not 
actually be related.

**6.** Check phenotype of cells

```{r}
#| label: fig-12
#| fig-cap: Proportions of phenotypes at time just after division (daughters), time just before division (parents), and all times.

#Set 1
a <- ggplot(daughters_df1) + 
  geom_bar(aes(Classifier.Phenotype), fill = "darkred") + 
  theme_light() + 
  xlab("class") + 
  ggtitle("Daughters") +
  ylim(c(0,3250))

b <- ggplot(parents_max1) + 
  geom_bar(aes(Classifier.Phenotype), fill = "darkblue") + 
  theme_light() + 
  xlab("class") + 
  ggtitle("Parents") +
  ylim(c(0,1750))

c <- ggplot(merged1) + 
  geom_bar(aes(Classifier.Phenotype)) + 
  theme_light() + 
  xlab("class") + 
  ggtitle("All Cells All Times") +
  ylim(c(0,30000))

#Set 2
d <- ggplot(daughters_df2) + 
  geom_bar(aes(Classifier.Phenotype), fill = "darkred") + 
  theme_light() + 
  xlab("class") + 
  ggtitle("Daughters") +
  ylim(c(0,3250))

e <- ggplot(parents_max2) + 
  geom_bar(aes(Classifier.Phenotype), fill = "darkblue") + 
  theme_light() + 
  xlab("class") + 
  ggtitle("Parents") +
  ylim(c(0,1750))

f <- ggplot(merged2) + 
  geom_bar(aes(Classifier.Phenotype)) + 
  theme_light() + 
  xlab("class") + 
  ggtitle("All Cells All Times") +
  ylim(c(0,30000))

(a+b+c)/(d+e+f)
```
It is important to see that parent cells are being appropriately identified as 
alive, as dead cells do not divide. It is possible for daughter cells to die 
following division, but it is concerning that the proportions of these match 
that of the parents and the general population. The top plots are Set 1, and 
the bottom plots are Set 2.
```{r}
#| warning: false
#| message: false

#Create a "is not in" function
'%!in%' <- function(x,y)!('%in%'(x,y))

#Find end point class
df <- filter(merged1, trackId != -1)
cells_list1 <- unique(df$trackId)
allcells_max1 <- data.frame()
for(i in cells_list1) {
  p <- filter(merged1, trackId == i)
  q <- max(p$t)
  r <- filter(p, t == q)
  if(i %in% parents_list1) {
    r <- mutate(r, parent = TRUE)
    allcells_max1 <- rbind(allcells_max1, r)
  }
  if(i %!in% parents_list1) {
    r <- mutate(r, parent = FALSE)
    allcells_max1 <- rbind(allcells_max1, r)    
  }
}

#Visualize class of cells at end points
allcells_max1 <- filter(allcells_max1, trackId != -1)
k <- ggplot(allcells_max1) + 
  geom_bar(aes(Classifier.Phenotype, fill = parent), position = 'dodge') + 
  theme_light() + 
  xlab("class") + 
  ggtitle("Class of Cells at Last Time Point")

#Get list of all time points labeled with dividing and parent statuses
#Time consuming step
allcells1 <- data.frame()
for(i in cells_list1) {
  p <- filter(merged1, trackId == i)
  q <- max(p$t)
  p <- cbind(p, lifetime = c(1:nrow(p)))
  r <- filter(p, t == q)
  s <- filter(p, t != q) %>% mutate(dividing = FALSE)
  if(i %in% parents_list1) {
    r <- mutate(r, dividing = TRUE)
    r <- rbind(s, r)
    r <- mutate(r, parent = TRUE)
    allcells1 <- rbind(allcells1, r)
  }
  if(i %!in% parents_list1) {
    r <- mutate(r, dividing = FALSE)
    r <- rbind(s, r)
    r <- mutate(r, parent = FALSE)
    allcells1 <- rbind(allcells1, r)    
  }
}

#Visualize divisions
merged_grouped1 <- merged1 %>% 
  filter(trackId != -1) %>% 
  group_by(t) %>% 
  summarize(class = mean(Class))

a <- ggplot(merged_grouped1) + 
  geom_col(aes(t, class)) + 
  ylab("class") + 
  xlab("time") + 
  ggtitle("Average Phenotype over Time") + 
  theme_light()

b <- ggplot(filter(allcells1, Class == 1)) + 
  geom_bar(aes(t, fill = dividing)) + 
  xlab("time") +
  ggtitle("Proportion of Live and Dividing Cells") + 
  theme_light()

c <- ggplot(allcells1) + 
  geom_bar(aes(t, fill = Classifier.Phenotype)) + 
  theme_light() + 
  xlab("time") + 
  labs(fill = "class") +
  ggtitle("Cell Phenotypes Over Time")

dividing_cells1 <- allcells1 %>% 
  filter(dividing == TRUE) 

d <- ggplot(dividing_cells1) + 
  geom_bar(aes(t, fill = Classifier.Phenotype)) + 
  theme_light() + 
  xlab("time") + 
  labs(fill = "class") +
  ggtitle("Divisions over Time")


#Set 2
df <- filter(merged2, trackId != -1)
cells_list2 <- unique(df$trackId)
allcells_max2 <- data.frame()
for(i in cells_list2) {
  p <- filter(merged2, trackId == i)
  q <- max(p$t)
  r <- filter(p, t == q)
  if(i %in% parents_list2) {
    r <- mutate(r, parent = TRUE)
    allcells_max2 <- rbind(allcells_max2, r)
  }
  if(i %!in% parents_list2) {
    r <- mutate(r, parent = FALSE)
    allcells_max2 <- rbind(allcells_max2, r)    
  }
}

#Visualize class of cells at end points
allcells_max2 <- filter(allcells_max2, trackId != -1)
l <- ggplot(allcells_max2) + 
  geom_bar(aes(Classifier.Phenotype, fill = parent), position = 'dodge') + 
  theme_light() + 
  xlab("class") + 
  ggtitle("Class of Cells at Last Time Point")

#Get list of all time points labeled with dividing and parent statuses
#Time consuming step
allcells2 <- data.frame()
for(i in cells_list2) {
  p <- filter(merged2, trackId == i)
  p <- cbind(p, lifetime = c(1:nrow(p)))
  q <- max(p$t)
  r <- filter(p, t == q)
  s <- filter(p, t != q) %>% mutate(dividing = FALSE)
  if(i %in% parents_list2) {
    r <- mutate(r, dividing = TRUE)
    r <- rbind(s, r)
    r <- mutate(r, parent = TRUE)
    allcells2 <- rbind(allcells2, r)
  }
  if(i %!in% parents_list2) {
    r <- mutate(r, dividing = FALSE)
    r <- rbind(s, r)
    r <- mutate(r, parent = FALSE)
    allcells2 <- rbind(allcells2, r)    
  }
}

#Visualize divisions
merged_grouped2 <- merged2 %>% 
  filter(trackId != -1) %>% 
  group_by(t) %>%
  summarize(class = mean(Class))

e <- ggplot(merged_grouped2) + 
  geom_col(aes(t, class)) + 
  ylab("class") + 
  xlab("time") + 
  ggtitle("Average Phenotype over Time") + 
  theme_light()

f <- ggplot(filter(allcells2, Class == 1)) + 
  geom_bar(aes(t, fill = dividing)) + 
  xlab("time") +
  ggtitle("Proportion of Live and Dividing Cells") + 
  theme_light()

g <- ggplot(allcells2) + 
  geom_bar(aes(t, fill = Classifier.Phenotype)) + 
  theme_light() + 
  xlab("time") + 
  labs(fill = "class") +
  ggtitle("Cell Phenotypes Over Time")

dividing_cells2 <- allcells2 %>% 
  filter(dividing == TRUE) 

h <- ggplot(dividing_cells2) + 
  geom_bar(aes(t, fill = Classifier.Phenotype)) + 
  theme_light() + 
  xlab("time") + 
  labs(fill = "class") +
  ggtitle("Divisions over Time")

k+l
a+e
b+f
c+g
d+h

#Visualize lifetime
a <- ggplot(allcells1) + 
  geom_histogram(aes(lifetime)) + 
  theme_light() + 
  ggtitle("Lifetime of Set 1")

b <- ggplot(allcells2) + 
  geom_histogram(aes(lifetime)) + 
  theme_light() + 
  ggtitle("Lifetime of Set 2")

a+b
```
These figures all point towards some important, general trends. The class of 
cells at their last time point is meant to show whether a cell is no longer 
being tracked because it has divided (alive, true) or because it has died 
(dead, false). A live cell at it's last time point, should it not be a parent,
would have lost it's tracking due to an error. A dead cell that is a parent has
been incorrectly classified as dead. 

The average phenotype graphs should remain the same, as they are based on the 
classifier data and not the tracking.

The other four graphs demonstrate the effects of the treatment. The left is 
Set 1 and the right is Set 2.

**6.** Ground truth-ish
```{r}
#| label: fig-15
#| fig-cap: The number of cells counted by hand and for each tracking set. Black line = identity line; blue line = regression
#| message: false
#| warning: false

a <- ggplot(filter(Book1, class == 'alive')) + 
  geom_bar(aes(slice)) + 
  scale_y_continuous(limits = c(0,1000)) + 
  theme_light() + xlab("t") + 
  ggtitle("Hand Annotation")

b <- ggplot(filter(allcells1, Class == 1)) + 
  geom_bar(aes(t)) + 
  scale_y_continuous(limits = c(0,1000)) + 
  theme_light() + 
  ggtitle("Set 1")

e <- ggplot(filter(allcells2, Class == 1)) + 
  geom_bar(aes(t)) + 
  scale_y_continuous(limits = c(0,1000)) + 
  theme_light() + 
  ggtitle("Set 2")

a+b+e

hand_ann <- Book1 %>% filter(class == "alive") %>% 
  group_by(slice) %>% 
  summarise(count = n())

b_ann <- allcells1 %>% filter(Class == 1) %>% 
  mutate(slice = t+1) %>% 
  group_by(slice) %>% 
  summarise(count_1 = n())

e_ann <- allcells2 %>% filter(Class == 1) %>% 
  mutate(slice = t+1) %>% 
  group_by(slice) %>% 
  summarise(count_2 = n())

abe <- left_join(hand_ann, b_ann)
abe <- left_join(abe, e_ann)
abe <- filter(abe, slice != 54)

p <- ggplot(abe) +
  geom_point(aes(count, count_1, color = slice)) +
  theme_light() +
  xlab("hand count") +
  ylab("set 1") +
  geom_smooth(aes(count, count_1), method = "lm", se = FALSE) +
  geom_abline(a = 0, b = 1) +
  ggtitle("Number of live cells tracked") +
  theme(legend.position = "none")
q <- ggplot(abe) +
  geom_point(aes(count, count_2, color = slice)) +
  theme_light() +
  geom_smooth(aes(count, count_2), method = "lm", se = FALSE) +
  geom_abline(a = 0, b = 1) +
  xlab("hand count") +
  ylab("set 2")

p+q


```


```{r}
#| label: fig-16
#| fig-cap: The number of cell divisions counted by hand and for each tracking set.
#| message: false
#| warning: false

c <- ggplot(Book2) + 
  geom_bar(aes(slice)) + 
  scale_y_continuous(limits = c(0,60)) + 
  theme_light() + 
  xlab("t") + 
  ggtitle("Hand Annotation") + 
  scale_x_continuous(limits = c(0,25))

d <- ggplot(dividing_cells1) + 
  geom_bar(aes(t)) + 
  scale_y_continuous(limits = c(0,60)) + 
  theme_light() + 
  ggtitle("Set 1")  + 
  scale_x_continuous(limits = c(0,25))

f <- ggplot(dividing_cells2) + 
  geom_bar(aes(t)) + 
  scale_y_continuous(limits = c(0,60)) + 
  theme_light() + 
  ggtitle("Set 2")  + 
  scale_x_continuous(limits = c(0,25))

c+d+f


hand_ann <- Book2 %>% 
  group_by(slice) %>% 
  summarise(count = n())

d_ann <- dividing_cells1 %>% 
  mutate(slice = t+1) %>% 
  group_by(slice) %>% 
  summarise(count_1 = n())

f_ann <- dividing_cells2 %>% 
  mutate(slice = t+1) %>% 
  group_by(slice) %>% 
  summarise(count_2 = n())

cdf <- left_join(hand_ann, d_ann)
cdf <- left_join(cdf, f_ann)

p <- ggplot(cdf) +
  geom_point(aes(count, count_1, color = slice)) +
  theme_light() +
  geom_smooth(aes(count, count_1), method = "lm", se = FALSE) +
  geom_abline(a = 0, b = 1) +
  xlab("hand count") +
  ylab("set 1") +
  ggtitle("Divisions tracked") +
  theme(legend.position = "none")
q <- ggplot(cdf) +
  geom_point(aes(count, count_2, color = slice)) +
  geom_smooth(aes(count, count_2), method = "lm", se = FALSE) +
  geom_abline(a = 0, b = 1) +
  theme_light() +
  xlab("hand count") +
  ylab("set 2")

p+q
```

**7.** Feature mapping
```{r}
#| label: fig-17
#| fig-cap: UMAP of cell features, Set 1.

#Note: the UMAPs have been "broken" for the most recent sets -> WIP

#Set 1
allcells1.data <- allcells1 %>% 
  drop_na() %>% 
  select(!c(1:6, 19:22, 39:42, 49:54, 56:69)) %>% 
  mutate(ID = row_number())
rownames(allcells1.data) <- NULL

allcells1.labels <- allcells1 %>% 
  drop_na %>% 
  select(!ID) %>% 
  mutate(ID = row_number())
rownames(allcells1.labels) <- NULL

umap_fit <- allcells1.data %>% 
  select(where(is.numeric)) %>% 
  column_to_rownames("ID") %>% 
  scale() %>% 
  umap()

umap_df <- umap_fit$layout %>% 
  as.data.frame() %>% 
  rename(UMAP1 = "V1", UMAP2 = "V2") %>% 
  mutate(ID = row_number()) %>%
  inner_join(allcells1.labels, by = "ID")

umap_df %>% 
  ggplot(aes(x = UMAP1, y = UMAP2, color = Class)) + 
  geom_point() + 
  labs(x = "UMAP1", y = "UMAP2", subtitle = "UMAP plot")
```
```{r}
#| label: fig-18
#| fig-cap: PCA of cell features, Set 1.

allcells1.data <- allcells1 %>% 
  drop_na() %>% 
  select(!c(t:mergerLabelId, starts_with("Center"), starts_with("Terminal"), 
            starts_with("Object_Center"), (ID:Class), dividing, parent, 
            Convex_Hull_Center_0, Convex_Hull_Center_1, 
            starts_with("Principal"), x, y, Size_in_pixels_0, 
            starts_with("Bounding_Box_Minimum"))) %>% 
  select(where(is.numeric)) %>% 
  scale()

corr <- cor(allcells1.data)
ggcorrplot(corr)

pca <- princomp(corr)
summary(pca)
fviz_eig(pca, addlabels = TRUE)
fviz_pca_var(pca, col.var = "cos2",
            gradient.cols = c("navy", "orchid", "palegreen"),
            repel = TRUE)
```
```{r}
#| label: fig-19
#| fig-cap: UMAP of cell features, Set 2.

#Set 2
allcells2.data <- allcells2 %>% drop_na() %>% 
  select(!c(t:mergerLabelId, starts_with("Center"), starts_with("Terminal"), 
            starts_with("Object_Center"), (ID:Class), dividing, parent, 
            Convex_Hull_Center_0, Convex_Hull_Center_1, 
            starts_with("Principal"), x, y, Size_in_pixels_0)) %>% 
  mutate(ID = row_number())
rownames(allcells2.data) <- NULL

allcells2.labels <- allcells2 %>% 
  drop_na %>% 
  select(!ID) %>% 
  mutate(ID = row_number())
rownames(allcells2.labels) <- NULL

umap_fit2 <- allcells2.data %>% 
  select(where(is.numeric)) %>% 
  column_to_rownames("ID") %>% 
  scale() %>% 
  umap::umap()

umap_df2 <- umap_fit2$layout %>% 
  as.data.frame() %>% 
  rename(UMAP1 = "V1", UMAP2 = "V2") %>% 
  mutate(ID = row_number()) %>%
  inner_join(allcells2.labels, by = "ID")

umap_df2 %>% 
  ggplot(aes(x = UMAP1, y = UMAP2, color = Class)) + 
  geom_point() + 
  labs(x = "UMAP1", y = "UMAP2", subtitle = "UMAP plot")
```
```{r}
#| label: fig-20
#| fig-cap: PCA of cell features, Set 2.

allcells2.data <- allcells2 %>% 
  drop_na() %>% 
  select(!c(t:mergerLabelId, starts_with("Center"), starts_with("Terminal"), 
            starts_with("Object_Center"), (ID:Class), dividing, parent, 
            Convex_Hull_Center_0, Convex_Hull_Center_1, 
            starts_with("Principal"), x, y, Size_in_pixels_0, 
            starts_with("Bounding_Box_Minimum"))) %>% 
  select(where(is.numeric)) %>% 
  scale()
  
corr <- cor(allcells2.data)
ggcorrplot(corr)

pca <- princomp(corr)
summary(pca)
fviz_eig(pca, addlabels = TRUE)
fviz_pca_var(pca, col.var = "cos2",
            gradient.cols = c("navy", "orchid", "palegreen"),
            repel = TRUE)
```

**8.** Distances
```{r}
#| label: fig-21
#| fig-cap: Mahalanobis distance by phenotype.
#| message: false
#| warning: false

#set 1
allcells1.data <- allcells1 %>% 
  drop_na() %>% 
  select(!c(t:mergerLabelId, starts_with("Center"),
            starts_with("Terminal"), Number_of_Holes_0,
            starts_with("Object_Center"), (ID:Class), dividing, parent, 
            Convex_Hull_Center_0, Convex_Hull_Center_1, lifetime,
            starts_with("Principal"), x, y, Size_in_pixels_0)) %>% 
  mutate(ID = row_number())

mal.1 <- allcells1.data
mal.1$mal <- mahalanobis(mal.1, colMeans(mal.1), cov(mal.1))
mal.1 <- inner_join(mal.1, allcells1.labels)

a <- ggplot(mal.1) + 
  geom_line(aes(t, mal, color = as.character(trackId))) + 
  theme_light() +
  theme(legend.position = "none") +
  ggtitle("Set 1")

#Set 2
allcells2.data <- allcells2 %>% 
  drop_na() %>% 
  select(!c(t:mergerLabelId, starts_with("Center"),
            starts_with("Terminal"), Number_of_Holes_0,
            starts_with("Object_Center"), (ID:Class), dividing, parent, 
            Convex_Hull_Center_0, Convex_Hull_Center_1, lifetime,
            starts_with("Principal"), x, y, Size_in_pixels_0)) %>% 
  mutate(ID = row_number())

mal.2 <- allcells2.data
mal.2$mal <- mahalanobis(mal.2, colMeans(mal.2), cov(mal.2))
mal.2 <- inner_join(mal.2, allcells2.labels)

b <- ggplot(mal.2) + 
  geom_line(aes(t, mal, color = as.character(trackId))) + 
  theme_light() +
  theme(legend.position = "none") +
  ggtitle("Set 2")

a+b

```
```{r}
#| label: fig-22
#| fig-cap: Euclidean distance by cell for Set 1.
#| message: false
#| warning: false

allcells1.data <- allcells1 %>% 
  drop_na() %>% 
  select(!c(t, labelimageId, lineageId:mergerLabelId, starts_with("Center"),
            starts_with("Terminal"), Number_of_Holes_0,
            starts_with("Object_Center"), (ID:Class), dividing, parent, 
            Convex_Hull_Center_0, Convex_Hull_Center_1, lifetime,
            starts_with("Principal"), x, y, Size_in_pixels_0)) %>% 
  mutate(ID = row_number())

#Most time consuming step; approx 30 minutes
cols <- data.frame()
data <- data.frame()
for(cell in unique(allcells1.data$trackId)) {
  df <- filter(allcells1.data, trackId == cell)
  means <- df %>% summarise_all(mean)
  cols <- data.frame(temp = c(1:nrow(df)))
  for(column in colnames(df)){
    if(column != "trackId") {
      col <- select(df, column)
      name <- colnames(col)
      value <- means[,column]
      col <- transmute(col, math = value - col[,1])
      colnames(col)[colnames(col) == "math"] = name
      cols <- cbind(cols, col)
    }
  }
  cols <- mutate(cols, trackId = cell)
  data <- rbind(data, cols)
}
data_b <- data
scaled.data <- scale(data[2:34])
scaled.data <- cbind(select(data, temp, trackId), scaled.data)
scaled.data$row_mean <- rowMeans(scaled.data[,c(3:35)], na.rm=TRUE)

a <- ggplot(scaled.data) + 
  geom_line(aes(temp, row_mean, color = as.character(trackId))) +
  theme_light() +
  theme(legend.position = "none") +
  xlab("lifetime") +
  ggtitle("Distance from cell mean over lifetime")

scaled.data <- rename(scaled.data, lifetime = temp)
scaled.data <- scaled.data %>% select(lifetime, trackId, row_mean)
allcells1_b <- allcells1

ts.data <- left_join(allcells1, scaled.data, by = c("lifetime", "trackId"))

b <- ggplot(ts.data) + 
  geom_line(aes(t, row_mean, color = as.character(trackId))) +
  theme_light() +
  theme(legend.position = "none") +
  ggtitle("Distance from cell mean over time")

a+b
```
```{r}
#| label: fig-23
#| fig-cap: Euclidean distance by cell for Set 2.
#| message: false
#| warning: false

allcells2.data <- allcells2 %>% 
  drop_na() %>% 
  select(!c(t, labelimageId, lineageId:mergerLabelId, starts_with("Center"),
            starts_with("Terminal"), Number_of_Holes_0,
            starts_with("Object_Center"), (ID:Class), dividing, parent, 
            Convex_Hull_Center_0, Convex_Hull_Center_1, lifetime,
            starts_with("Principal"), x, y, Size_in_pixels_0)) %>% 
  mutate(ID = row_number())

#Most time consuming step; approx 30 minutes
cols <- data.frame()
data.2 <- data.frame()
for(cell in unique(allcells2.data$trackId)) {
  df <- filter(allcells2.data, trackId == cell)
  means <- df %>% summarise_all(mean)
  cols <- data.frame(temp = c(1:nrow(df)))
  for(column in colnames(df)){
    if(column != "trackId") {
      col <- select(df, column)
      name <- colnames(col)
      value <- means[,column]
      col <- transmute(col, math = value - col[,1])
      colnames(col)[colnames(col) == "math"] = name
      cols <- cbind(cols, col)
    }
  }
  cols <- mutate(cols, trackId = cell)
  data.2 <- rbind(data.2, cols)
}
data2_b <- data.2
scaled.data2 <- scale(data.2[2:34])
scaled.data2 <- cbind(select(data.2, temp, trackId), scaled.data2)
scaled.data2$row_mean <- rowMeans(scaled.data2[,c(3:35)], na.rm=TRUE)

a <- ggplot(scaled.data2) + 
  geom_line(aes(temp, row_mean, color = as.character(trackId))) +
  theme_light() +
  theme(legend.position = "none") +
  xlab("lifetime") +
  ggtitle("Distance from cell mean over lifetime")

scaled.data2 <- rename(scaled.data2, lifetime = temp)
scaled.data2 <- scaled.data2 %>% select(lifetime, trackId, row_mean)
allcells2_b <- allcells2

ts.data2 <- left_join(allcells2, scaled.data2, by = c("lifetime", "trackId"))

b <- ggplot(ts.data2) + 
  geom_line(aes(t, row_mean, color = as.character(trackId))) +
  theme_light() +
  theme(legend.position = "none") +
  ggtitle("Distance from cell mean over time")

a+b
```

**9.** Time series features

```{r}
#| label: fig-24
#| fig-cap: Time series features for Set 1.
#| message: false
#| warning: false

ts.filt <- ts.data %>% select(t, trackId, row_mean) %>% drop_na() %>% 
  group_by(trackId)

ts.filt$t <- as_date(ts.filt$t)

ts.xts <- ts.filt %>%
         group_by(trackId) %>%
         pivot_wider(names_from = "trackId", values_from = "row_mean", 
         values_fill = list(row_mean = 0))

names <- colnames(ts.xts)

ts_xts <- xts(ts.xts[-1], as.Date(ts.xts$t, "%d"))
tsxts_feat <- tsfeatures(ts_xts)

a <- ggplot(tsxts_feat) +
  geom_histogram(aes(trend)) +
  theme_light()
b <- ggplot(tsxts_feat) +
  geom_histogram(aes(spike)) +
  theme_light()
c <- ggplot(tsxts_feat) +
  geom_histogram(aes(linearity)) +
  theme_light()
d <- ggplot(tsxts_feat) +
  geom_histogram(aes(curvature)) +
  theme_light()
e <- ggplot(tsxts_feat) +
  geom_histogram(aes(e_acf1)) +
  theme_light()
f <- ggplot(tsxts_feat) +
  geom_histogram(aes(e_acf10)) +
  theme_light()
g <- ggplot(tsxts_feat) +
  geom_histogram(aes(entropy)) +
  theme_light()
h <- ggplot(tsxts_feat) +
  geom_histogram(aes(x_acf1)) +
  theme_light()
i <- ggplot(tsxts_feat) +
  geom_histogram(aes(x_acf10)) +
  theme_light()
j <- ggplot(tsxts_feat) +
  geom_histogram(aes(diff1_acf1)) +
  theme_light()
k <- ggplot(tsxts_feat) +
  geom_histogram(aes(diff1_acf10)) +
  theme_light()
l <- ggplot(tsxts_feat) +
  geom_histogram(aes(diff2_acf1)) +
  theme_light()
m <- ggplot(tsxts_feat) +
  geom_histogram(aes(diff2_acf10)) +
  theme_light()

(a+b+c+d)/(e+f+g)
(h+i+j)/(k+l+m)
plot(tsxts_feat$entropy, tsxts_feat$spike)
```

```{r}
#| label: fig-25
#| fig-cap: Time series features UMAP for Set 1.
#| message: false
#| warning: false

tsxts_feat.data <- tsxts_feat %>% 
  select(!c(frequency:seasonal_period)) %>% 
  mutate(ID = row_number()) %>% 
  drop_na()

tsxts_feat.labels <- tsxts_feat %>% 
  mutate(ID = row_number()) %>% 
  drop_na

umap_fit_ts <- tsxts_feat.data %>% 
  select(where(is.numeric)) %>% 
  column_to_rownames("ID") %>% 
  scale() %>% 
  umap()

umap_df_ts <- umap_fit_ts$layout %>% 
  as.data.frame() %>% 
  rename(UMAP1 = "V1", UMAP2 = "V2") %>% 
  mutate(ID = row_number()) %>%
  inner_join(tsxts_feat.labels, by = "ID")

umap_df_ts %>% 
  ggplot(aes(x = UMAP1, y = UMAP2, color = ID)) + 
  geom_point() + 
  labs(x = "UMAP1", y = "UMAP2", subtitle = "UMAP plot")
```
```{r}
#| label: fig-26
#| fig-cap: Time series features for Set 2.
#| message: false
#| warning: false

ts.filt2 <- ts.data2 %>% select(t, trackId, row_mean) %>% drop_na() %>% 
  group_by(trackId)

ts.filt2$t <- as_date(ts.filt2$t)

ts.xts2 <- ts.filt2 %>%
         group_by(trackId) %>%
         pivot_wider(names_from = "trackId", values_from = "row_mean", 
         values_fill = list(row_mean = 0))
names2 <- colnames(ts.xts2)

ts_xts2 <- xts(ts.xts2[-1], as.Date(ts.xts2$t, "%d"))
tsxts_feat2 <- tsfeatures(ts_xts2)

a <- ggplot(tsxts_feat2) +
  geom_histogram(aes(trend)) +
  theme_light()
b <- ggplot(tsxts_feat2) +
  geom_histogram(aes(spike)) +
  theme_light()
c <- ggplot(tsxts_feat2) +
  geom_histogram(aes(linearity)) +
  theme_light()
d <- ggplot(tsxts_feat2) +
  geom_histogram(aes(curvature)) +
  theme_light()
e <- ggplot(tsxts_feat2) +
  geom_histogram(aes(e_acf1)) +
  theme_light()
f <- ggplot(tsxts_feat2) +
  geom_histogram(aes(e_acf10)) +
  theme_light()
g <- ggplot(tsxts_feat2) +
  geom_histogram(aes(entropy)) +
  theme_light()
h <- ggplot(tsxts_feat2) +
  geom_histogram(aes(x_acf1)) +
  theme_light()
i <- ggplot(tsxts_feat2) +
  geom_histogram(aes(x_acf10)) +
  theme_light()
j <- ggplot(tsxts_feat2) +
  geom_histogram(aes(diff1_acf1)) +
  theme_light()
k <- ggplot(tsxts_feat2) +
  geom_histogram(aes(diff1_acf10)) +
  theme_light()
l <- ggplot(tsxts_feat2) +
  geom_histogram(aes(diff2_acf1)) +
  theme_light()
m <- ggplot(tsxts_feat2) +
  geom_histogram(aes(diff2_acf10)) +
  theme_light()

(a+b+c+d)/(e+f+g)
(h+i+j)/(k+l+m)
plot(tsxts_feat$entropy, tsxts_feat$spike)
```
The trend is the overall movement of the time series stripped of its periodic 
behaviors. It is meant to trace the long term pattern of a time series, and it
can be thought of in terms of general increasing or decreasing. 

The spike measures the “spikiness” of a time series, and is computed as the 
variance of the leave-one-out variances of the remainder component. The 
remainder is what remains once the trend and periodic behaviors have been 
removed from the data. This would (ideally) be composed of random spikes.

Linearity and curvature of a time series calculated based on the coefficients 
of an orthogonal quadratic regression. This form of regression is less sensitive
to oscillation (which is often produced by periodic behaviors).

Features containing "acf":
The autocorrelation function of the series, the differenced series, and the 
twice-differenced series. The autocorrelation function finds the correlation 
between a time series and itself one time period apart (lag 1). A high 
correlation means that previous time points are more able to be used to predict 
later time points.

Entropy measures the “forecastability” of a time series, where low values 
indicate a high signal-to-noise ratio, and large values occur when a series is 
difficult to forecast.nm

```{r}
#| label: fig-27
#| fig-cap: Time series features UMAP for Set 2.
#| message: false
#| warning: false

tsxts_feat.data <- tsxts_feat2 %>% 
  select(!c(frequency:seasonal_period)) %>% 
  mutate(ID = row_number()) %>% 
  drop_na()

tsxts_feat.labels <- tsxts_feat2 %>% 
  mutate(ID = row_number()) %>% 
  drop_na

umap_fit_ts <- tsxts_feat.data %>% 
  select(where(is.numeric)) %>% 
  column_to_rownames("ID") %>% 
  scale() %>% 
  umap()

umap_df_ts <- umap_fit_ts$layout %>% 
  as.data.frame() %>% 
  rename(UMAP1 = "V1", UMAP2 = "V2") %>% 
  mutate(ID = row_number()) %>%
  inner_join(tsxts_feat.labels, by = "ID")

umap_df_ts %>% 
  ggplot(aes(x = UMAP1, y = UMAP2, color = ID)) + 
  geom_point() + 
  labs(x = "UMAP1", y = "UMAP2", subtitle = "UMAP plot")
```

## Tracking visualization
**1.** Showing all cells over time, colored by days alive. 
```{r}
#Sets 1 & 2
palette(rainbow(53))
fi <- list.files(paste0(root, filesep, "B01_20230407_Incucyte_Images_woGFP_Analysis_QI_Core"),pattern=well,full.names = T)
fi <- fi[2:54]
merged=list(merged1=merged1, merged2=merged2)
for(set in names(merged)){
  for(i in unique(allcells2$t)){
    p <- allcells2 %>% filter(t == i, trackId != -1)
    tiff(paste0("~/Downloads/track", merged[[set]][1,"ID"], "_", i, ".tif"), width = 1408, height = 1040)
    img=bioimagetools::readTIF(fi[i+1],as.is = T)
    img <- EBImage::resize(img, dim(img)[1]/1)
    plot(raster::as.raster(img[,,,1]))
    points(p$x, p$y, pch=(p$Class+2), col = p$lifetime, cex=1.5, lwd = 2)
    text(p$x+20, p$y, labels=p$trackId, cex=1)
    mtext(fileparts(fi[i+1])$name)
    legend("topleft", legend = unique(allcells2$Classifier.Phenotype), pch=unique(allcells2$Class+2), cex = 1)
    dev.off()
  }
}

```
**2.** Show one track at a time.
```{r}
## visualize one track at a time on raw images
# fi=list.files(paste0(root, filesep, "B01_20230407_Incucyte_Images_woGFP_Analysis_QI_Core"),pattern=well,full.names = T)
# fi=list.files(paste0(root, filesep, "I01_20231219_HALOCellClassification/E2_1_Registered_image"),pattern=well,full.names = T)
fi=list.files(paste0(root, filesep, "I01_20231219_HALOCellClassification/E2_1_Overlays"),pattern=well,full.names = T)
# trackIDs=names(la)[order(sapply(la,nrow),decreasing = T)]
trackIDs=as.character(order(tsxts_feat$linearity,decreasing = T))
trackIDs=trackIDs[trackIDs %in% names(la1)]
# goodTracks = c("136")
# match(goodTracks,trackIDs)
# badTracks = c("1057")
# match(badTracks,trackIDs)
for(trackID in c(head(trackIDs),tail(trackIDs))){
  toi=la1[[trackID]]
  pdf(paste0("~/Downloads/testTrack_",trackID,".pdf"), width=14, height = 9)
  
  plot(toi$t,toi$Object_Area_0, col=toi$Class+2,pch=20); 
  legend("topleft",unique(toi$Classifier.Phenotype), fill=unique(toi$Class+2))
  
  fidx=max(2,grep(toi$time[1],fi,value=F))
  center=NULL
  # for(i in 1:nrow(toi)){
  while(fidx<=length(fi)){  
    thistime=strsplit(fileparts(fi[fidx])$name,"_")[[1]][4]
    i=which(toi$time==thistime)
    if(is.null(center)){
      center=c(toi$x[i], toi$y[i])
    }
    img=bioimagetools::readTIF(fi[fidx],as.is = T)
    img <- EBImage::resize(img, dim(img)[1]/1)
    par(mfrow=c(1,2))
    plot(raster::as.raster(img[,,,1]), xlim=c(center[1]-200,center[1]+200), ylim=c(center[2]-50,center[2]+50))
    mtext(fileparts(fi[fidx])$name)
    ## Just for testing
    # tmp=merged[merged$t==toi$t[i],]
    # points(tmp$x,tmp$y,pch=3,col="green")
    if(!isempty(i)){
      points(toi$x[i], toi$y[i],pch=3,cex=3)
    }
    plot(toi$t,toi$Object_Area_0, col=c("black","red")[1+(toi$time==thistime)],pch=20,cex=2); 
    fidx=fidx+1;
  }
  dev.off()
}
```
**3.** Show overlays and tracking ID.
```{r}
## plot overlays along with tracking ID:
pdf(paste0("~/Downloads/trackIDs.pdf"))
fidx=2
while(fidx<=length(fi)){  
  img=bioimagetools::readTIF(fi[fidx],as.is = T)
  img <- EBImage::resize(img, dim(img)[1]/1)
  plot(raster::as.raster(img[,,,1]), xlim=c(211,700), ylim=c(150,650))
  mtext(fileparts(fi[fidx])$name)
  
  thistime=strsplit(fileparts(fi[fidx])$name,"_")[[1]][4]
  tmp=sapply(names(la1), function(x) la1[[x]][la1[[x]]$time==thistime,,drop=F], simplify = F)
  tmp=sapply(names(tmp), function(x) cbind(tmp[[x]], rep(x, nrow(tmp[[x]]))), simplify = F)
  tmp=do.call(rbind, tmp)
  tmp=as.data.frame(tmp)
  tmp=tmp[which(!sapply(tmp$x,isempty)),]
  colnames(tmp)[ncol(tmp)]="lineageID"
  text(unlist(tmp$x), unlist(tmp$y),labels = unlist(tmp$lineageID), cex=0.6)
  fidx=fidx+1;
}
dev.off()
